---
number-sections: false
toc-depth: 3
unnumbered: true
---

# Preface {.unnumbered}

This workshop is a hands-on introduction to writing reproducible, publication-ready research documents in R and Quarto. Across two sessions, we'll cover the full workflow from raw data to a formatted manuscript, including summary tables, model outputs, figures, citations, and export to HTML, PDF, Word and LaTeX.

## Who this is for

Researchers with a working knowledge of R and the tidyverse who want to move from exploratory analysis to polished, reproducible manuscripts.

## What you will learn

By the end of this workshop, you will be able to:

- Write and structure documents in Quarto using Markdown
- Export the same document to multiple formats (HTML, Word, PDF, LaTeX) with minimal format-specific tweaking
- Incorporate data directly into your text using inline reporting
- Produce publication-ready summary tables and model output tables
- Create and export figures with `ggplot2` and cross-reference them in your manuscript
- Manage code chunk options to control what appears in your output

## Sessions

**Day 1** focuses on the foundations: writing in Quarto, producing different 
output formats from a single source document, and bringing your data into the 
narrative through summary tables and inline reporting. This will be broken into the following chapters:

- Markdown basics: writing in Quarto
- Output formats: HTML, Word, PDF, and LaTeX
- Incorporating data: summary tables and inline reporting

**Day 2** builds on this to cover the statistical reporting workflow: 
extracting and presenting model results, producing publication-ready figures, 
and tying everything together with cross-references. This will be broken into the following chapters:

- Reporting model results: model summary tables and inline reporting
- Producing figures: publication-ready plots, exporting, and cross-referencing

Tips on code chunk options are woven throughout both days.

## How to use this book

Each chapter corresponds to roughly 30 minutes of workshop time. Code chunks are fully executable, work through them in your own RProject alongside the book. Each chapter also has accompanying slides (linked at the top) for the taught portion.

## Required software

Before the first session, please make sure you have the following installed:

- **R** ($\geq$ 4.2): <https://cran.r-project.org>
- **RStudio** ($\geq$ 2022.07): <https://posit.co/download/rstudio-desktop>
- **Quarto** (latest): <https://quarto.org/docs/download>
- **TinyTeX** (for PDF output) — see [Installing TinyTeX] below

### Installing R, RStudio, and Quarto

Download and install each from the links above, following the instructions 
for your operating system. Install them in order: R first, then RStudio, 
then Quarto.

### Installing TinyTeX

TinyTeX is a lightweight LaTeX distribution installable directly from R:

```{r}
#| eval: false
install.packages("tinytex")
tinytex::install_tinytex()
```

### Checking your versions

In the R console:
```{r}
#| output: false
R.version.string
tinytex::is_tinytex()
tinytex::tlmgr("--version")
```

In the **Terminal** tab:
```bash
quarto --version
```

RStudio's version is visible under *Help → About RStudio*.

### Updating

- **R**: download the latest version from <https://cran.r-project.org>
- **RStudio**: *Help → Check for Updates*
- **Quarto**: download the latest version from <https://quarto.org/docs/download>
- **TinyTeX**: `tinytex::tlmgr_update()`

### Packages

We'll be using the `pacman` package to load and install required (CRAN) packages. As long as you have this package installed you should be able to run the example and exercise code.

```{r}
#| eval: false
install.packages("pacman")
```

You only need to install `pacman` once. After that, `pacman::p_load()` handles loading and installing any missing packages automatically. Avoid leaving `install.packages()` calls uncommented in your scripts as they will re-run every time the script is executed or the document is rendered, which is slow and can cause unexpected package version changes. In other words, run this in the Console, or comment it out if you have it in a script.

## Disclaimer {.unnumbered}

These materials were developed with the assistance of an AI language model (Claude, Anthropic) for outlining and editorial refinement using guided prompts. All content was fully reviewed, improved (I hope), and edited by the author. Every topic and reference included falls within the author's own expertise; the use of AI assistance was a practical decision about time and effort, not a substitution for subject knowledge. Any errors, inaccuracies, or omissions are the author's own responsibility.

These chapters also lean heavily on materials I previously created (without AI assistance) on R for Reproducibility, both in a semester-length course [	Open Science Practices: Implementing a Reproducible Analysis Workflow](https://daniela-palleschi.github.io/r4repro_SoSe2024/), a 4-day workshop [Open Science Practices for Linguistic Research: Reproducible Analyses in R](https://daniela-palleschi.github.io/SSOL24-reproducibility-workshop/), and a 2-part workshop [Reproducible Workflow in R](https://daniela-palleschi.github.io/zas-reproducibility-2024/). Unlike these previous materials which covered reproducibility and dynamic reports more broadly (and more in-depth), this book focuses on dynamic reports and reproducible analyses in the context of reporting results for publication and writing a manuscript in general.


