---
title: "Summary Statistics & Demographics Tables"
subtitle: "Tables and in-line reporting"
---

```{r}
#| output: false
pacman::p_load(
  here,
  tidyverse,
  gtsummary,
  gt)
```

```{r}
#| echo: true
df_clean <- read_csv(here::here("data", "processed", "exp1_clean.csv"))
```

## The demographics table

A demographics or "Table 1" is usually the first table in a methods section. It describes the sample: who was measured, on what variables, and how those break down by group.

## Summary tables

`gtsummary` produces publication-ready tables from data frames with minimal code and sensible defaults.

### Basic summary table

#### Total reading time

```{r}
sum_tt <-
  df_clean |>
  filter(region == "verb") |>
  select(tense, lifetime, tt) |>
  mutate(
    lifetime = factor(lifetime, levels = c("living", "dead"))
  ) |>
  summarise(
    n      = n(),
    mean   = mean(tt,   na.rm = TRUE),
    sd     = sd(tt,     na.rm = TRUE),
    median = median(tt, na.rm = TRUE),
    iqr    = IQR(tt,    na.rm = TRUE), # inter-quartile range (middle 50% of data)
    se     = sd(tt, na.rm = TRUE) / sqrt(sum(!is.na(tt))),
    ci     = qt(0.975, df = sum(!is.na(tt)) - 1) * sd(tt, na.rm = TRUE) / sqrt(sum(!is.na(tt))),
    .by    = c(tense, lifetime)
  )
```


```{r}
sum_tt |> 
  gt() |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  cols_label(
    tense    = "Tense",
    lifetime = "Lifetime",
    n        = "N",
    mean     = "Mean",
    sd       = "SD",
    median   = "Median",
    iqr      = "IQR",
    se       = "SE",
    ci       = "95% CI"
  ) |>
  tab_spanner(
    label   = "Total reading time (ms)",
    columns = c(mean, sd, median, iqr, se, ci)
  ) |>
  fmt_number(columns = c(mean, sd, median, iqr, se, ci), decimals = 1)
```

::: callout-note
`add_p()` automatically selects tests (t-test, Wilcoxon, chi-square) based on variable type. You can override with `test = list(...)`.
:::

### Stratified by multiple groups

```{r}
df_clean |>
  select(species, sex, mass) |>
  tbl_strata(
    strata = species,
    .tbl_fun = ~ .x |>
      tbl_summary(
        by = sex,
        statistic = all_continuous() ~ "{mean} ({sd})",
        digits = all_continuous() ~ 1
      )
  )
```

## Exporting gtsummary tables

For Word:

```{r}
#| eval: false
tbl <- df_clean |>
  select(species, sex, mass) |>
  tbl_summary(by = sex) |>
  as_flex_table()

flextable::save_as_docx(tbl, path = here::here("output", "tables", "table1.docx"))
```

For LaTeX:

```{r}
#| eval: false
tbl |>
  as_kable_extra(format = "latex", booktabs = TRUE) |>
  kableExtra::save_kable(here::here("output", "tables", "table1.tex"))
```

## Chapter slides

```{=html}
<a href="../slides/03_demographics.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
```
