---
title: "Data and Code in Quarto"
subtitle: "Code Chunks, Loading Data, and Inline Reporting"
author: "Daniela Palleschi"
date: "2026-03-25"
date-format: "MMMM D, YYYY"
day: "Day 2"
---

## Overview

::: incremental
- Code chunks: structure and options
- Loading and saving data
- Inline reporting
:::

### The dataset

We will use eye-tracking reading data from @palleschi_living_2025, adapted for this workshop.

::: incremental
- **Design:** 2 × 2 — lifetime (living/dead) × tense (present perfect/simple future)
- **Regions:** `verb` of interest for today
- **Reading measures:** `gaze` (first-pass reading time) for today
- **Additional:** `rating` (1–7 plausibility), `rt_ms` (reaction time to rating)
- **Note:** participant codes (`px_code`) are randomly generated; we will anonymise these later
:::

. . .

::: callout-tip
The verb region is the most theoretically relevant. We will focus on `gaze` in most exercises.
:::

# Code chunks {background-color="#8D203D" data-stack-name="Code chunks"}

## Structure

````markdown
## Reaction times

Calculate mean reaction times:

```{{r}}
rt <- c(312, 287, 445, 301, 389)
mean(rt)
```
````


. . .

- Fenced with ` ```{r} ` and ` ``` `
- Executed in sequence; output inserted at that position
- Code must be written linearly: top to bottom

## Chunk options

:::: columns
::: {.column width="50%"}
````markdown
```{{r}}
#| echo: false
#| warning: false
5 * 6
```
````

- must appear *directly* below ` ```{r} `, and
- be preceded with `#|` + an empty space, and
- be followed by a colon + empty space + value
- apply only to when rendering or 'Run all chunks (above)', not when running interactively
:::
::: {.column width="50%"}
::: {style="font-size: 0.7em;"}
| Option | Effect |
|---|---|
| `echo` | show or hide code |
| `output` | show or hide output |
| `eval` | run or skip chunk |
| `include` | include chunk and output |
| `warning` | show or hide warnings |
| `message` | show or hide messages |
| `error` | halt on error (`false`) or continue and display error (`true`) |

: {tbl-colwidths="[20,80]"}
:::
:::
::::

## Global defaults and overrides

Set globally in YAML:
````yaml
execute:
  echo: false
  warning: false
  message: false
````

. . .

Override per chunk:

```{r}
#| echo: fenced
#| eval: true
#| error: false
2+2
```

::: {style="text-align: center; margin-top: 1em; color: white;"}
### ✏️ Exercises — 5 minutes {background-color="#8D203D"}

[Exercises 1 - 3: Code chunks](file:///C:/Users/Palleschi/OneDrive%20-%20Leibniz-ZAS/Documents/Workshops/publishr/_book/exercises/day2.html#data-and-code)
:::


# Loading data {background-color="#8D203D" data-stack-name="Loading data"}

## Packages

Always load pacakges at the top of your script!

```{r}
#| echo: fenced
library(here)
library(tidyverse)
```

I like the `pacman::p_load()` function for loading packages instead.

```{r}
#| echo: true
# install.packages("pacman")
pacman::p_load(here, tidyverse)
```


## here::here()

Always use `here::here()` for project-relative paths:

```{r}
#| echo: fenced
df       <- readr::read_csv(here::here("data", "raw", "data_example.csv"))
fit      <- readRDS(here::here("output", "models", "lmer_gaze.rds"))
```

. . .

Save processed objects so the manuscript never re-runs the analysis:

```{r}
#| echo: fenced
#| eval: false
readr::write_csv(df_clean, here::here("data", "processed", "data_clean.csv"))
```

. . .

::: callout-note
The **analysis** script **processes/analyses** data and **saves it** under a new filename. 

The **manuscript** script **loads in** the processed/analysed data and models and **reports** them.
:::

::: {style="text-align: center; margin-top: 1em; color: white;"}
### ✏️ `eval: true` or `false`? {background-color="#8D203D"}

In an analysis script, which code chunks should be evaluated when rendering?

- loading in your processed data
- wrangling or checking distributions
- running models
- saving models as `.rds` files
- re-loading in saved models
- printing model summaries
- running model diagnostics
:::


## Anonymising data

- here we create an anonymised participant code and save the result:

```{r}
#| eval: false
#| echo: true
# create anonymised participant codes
set.seed(416)

px_lookup <- df |>
  distinct(px_code) |> # for each distinct px_code
  mutate(px = sample(1:n(), n(), replace = FALSE)) # randomly assign a value of 1:totalN

df_clean <- df |>
  left_join(px_lookup, by = "px_code") |>
  select(-px_code)
```

## Saving cleaned data

- cleaning/processing usually includes a lot more than just anonymise your data
- for now let's leave it at that

```{r}
#| echo: fenced
#| eval: false
saveRDS(here("data", "processed", "data_clean.csv"))
```

- `eval: false` because we don't want it to be saved with each render
- usually at this point you would/should delete the raw data from your local machine
    + and *archive* and *encrypt* it somewhere secure
- note that we saved it in a subfolder for `processed` data

::: {style="text-align: center; margin-top: 1em; color: white;"}
### ✏️ Exercises — 5 minutes {background-color="#8D203D"}

[Exercises 4 - 6: Load data](https://daniela-palleschi.github.io/publishr/exercises/day1.html#exercise-4-load-data)
:::

# Inline reporting {background-color="#8D203D" data-stack-name="Inline reporting"}

## The idea

Instead of typing numbers manually:

> "The sample comprised **48** participants."

. . .

Embed directly in prose:
````markdown
The sample comprised ` r dplyr::n_distinct(df$px)` participants.
````

. . .

Re-render when data changes -- numbers update automatically.

## Store values as objects
```{r}
#| eval: false
n_px  <- dplyr::n_distinct(df_clean$px)
m_tt  <- round(mean(df_clean$tt, na.rm = TRUE), 1)
sd_tt <- round(sd(df_clean$tt,   na.rm = TRUE), 1)
```

. . .
````markdown
The sample comprised ` r n_px` participants.
Mean total reading time was ` r m_tt` ms (*SD* = ` r sd_tt`).
````

. . .

> The sample comprised 24 participants. Mean total reading time was 350 ms (*SD* = 56).


::: {style="text-align: center; margin-top: 1em; color: white;"}
### ✏️ Exercises — 10 minutes {background-color="#8D203D"}

[Exercises 7 - 9](https://daniela-palleschi.github.io/publishr/exercises/day1.html#exercise-5-summary-statistics)
:::



## Summary

::: incremental
- Chunk options control what is shown and what runs
- `here::here()` makes paths portable — the manuscript only loads, never re-runs
- Inline R keeps reported values in sync with your analysis
- Any number in your manuscript should be computed, not typed
:::

## Up next... {background-color="#8D203D"}

::: {style="text-align: center; margin-top: 2em; color: white;"}
### Reporting models

*Analyses and manuscripts in Quarto*
:::

# References

