---
title: "Figures for Publication"
subtitle: "ggplot2, themes, and exporting"
---

```{r}
#| output: false
library(tidyverse)
library(patchwork)
```

```{r}
#| output: false
df_clean <- palmerpenguins::penguins |>
  filter(!is.na(sex)) |>
  mutate(
    sex     = factor(sex, labels = c("Female", "Male")),
    species = factor(species)
  )
```

## A publication theme

Define a base theme once and apply it consistently. This makes all your figures coherent and easy to update.

```{r}
theme_pub <- function(base_size = 11) {
  theme_classic(base_size = base_size) +
    theme(
      strip.background  = element_blank(),
      strip.text        = element_text(face = "bold"),
      legend.position   = "bottom",
      legend.title      = element_blank(),
      axis.line         = element_line(colour = "grey40"),
      axis.ticks        = element_line(colour = "grey40"),
      panel.grid.major.y = element_line(colour = "grey92")
    )
}

theme_set(theme_pub())
```

## Common figure types

### Grouped means with error bars

```{r}
#| fig-cap: "Mean body mass (±1 SD) by species and sex."
#| label: fig-mass

df_clean |>
  filter(!is.na(body_mass_g)) |>
  group_by(species, sex) |>
  summarise(
    mean_mass = mean(body_mass_g),
    sd_mass   = sd(body_mass_g),
    .groups   = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = sex, group = sex)) +
  geom_point(position = position_dodge(0.4), size = 3) +
  geom_errorbar(
    aes(ymin = mean_mass - sd_mass, ymax = mean_mass + sd_mass),
    width = 0.2,
    position = position_dodge(0.4)
  ) +
  scale_colour_manual(values = c("Female" = "#C0392B", "Male" = "#2980B9")) +
  labs(x = NULL, y = "Body mass (g)")
```

### Distribution plot

```{r}
#| fig-cap: "Flipper length distributions by species."
#| label: fig-flipper

ggplot(df_clean, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(bins = 25, alpha = 0.7, position = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Flipper length (mm)", y = "Count")
```

### Combining figures with patchwork

```{r}
#| fig-cap: "Combined figure: mass by group (A) and flipper distribution (B)."
#| label: fig-combined
#| fig-width: 10
#| fig-height: 4

p1 <- df_clean |>
  filter(!is.na(body_mass_g)) |>
  group_by(species, sex) |>
  summarise(mean_mass = mean(body_mass_g), sd_mass = sd(body_mass_g), .groups = "drop") |>
  ggplot(aes(x = species, y = mean_mass, colour = sex, group = sex)) +
  geom_point(position = position_dodge(0.4), size = 3) +
  geom_errorbar(
    aes(ymin = mean_mass - sd_mass, ymax = mean_mass + sd_mass),
    width = 0.2, position = position_dodge(0.4)
  ) +
  scale_colour_manual(values = c("Female" = "#C0392B", "Male" = "#2980B9")) +
  labs(x = NULL, y = "Body mass (g)")

p2 <- ggplot(df_clean, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(bins = 25, alpha = 0.7, position = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Flipper length (mm)", y = "Count")

(p1 | p2) + plot_annotation(tag_levels = "A")
```

## Exporting figures

```{r}
#| eval: false
ggsave(
  here::here("output", "figures", "fig1_mass.pdf"),
  plot   = p1,
  width  = 8,
  height = 5,
  device = cairo_pdf  # best for fonts in LaTeX
)

ggsave(
  here::here("output", "figures", "fig1_mass.png"),
  plot  = p1,
  width = 8,
  height = 5,
  dpi   = 300          # 300 dpi for most journals
)
```

::: callout-tip
For LaTeX manuscripts, prefer `cairo_pdf` over `pdf` — it handles Unicode and custom fonts more reliably. For Word, use high-DPI PNG (300+).
:::

## Chapter slides

```{=html}
<a href="../slides/04_figures.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
```
