---
title: "Reporting Model Results"
subtitle: "modelsummary, broom, and inline reporting"
---

## Overview

::: incremental
- workflow for fitting a linear mixed effect model
- extracting estimates for inline reporting
- producing model summary table
:::

## Set-up

- load packages: `here`, `tidyverse`, `lme4`, `lmerTest`, `kableExtra`
- load data: `data/processed/data_clean.csv`; save it as `df_clean`
  + or load some data you want to work with
- load helper function for formatting p-values: `source(here("R", "helpers.R"))`

### Code

```{r}
#| eval: true
pacman::p_load(here, tidyverse, lme4, lmerTest, kableExtra)
df_clean <- read_csv(here("data", "processed", "data_clean.csv"))
source(here("R", "helpers.R"))
```

```{r}
#| echo: false
#| eval: false

# run this only when running interactively
pacman::p_load(here, tidyverse, lme4, lmerTest, kableExtra)
df_clean <- read_csv(here("../", "data", "processed", "data_clean.csv"))
```

## Model fitting

- even with pre-processed data you typically want to do some data wrangling and exploration
- usually I'd suggest having an Exploratory Data Analysis script before your analysis script
    + here you'd plot the data in all different ways to get a feel for it
    + you'd also want to look at the observations across participants etc.
- we'll skip this step

### Data prep

- we need to filter our data
- apply contrast coding for factorial predictors
- centre continuous predictors
- inspect distribution of dependent variable

### Fit model

- we'll do frequentist, and just run the published model structure

```{r}
#| echo: fenced
#| eval: false
lmer_gaze <-
  lmer(log(gaze) ~ lifetime * tense + trial_c +
         (1 + lifetime | px) +
         (1 + tense | item), 
       data    = df_verb,
       control = lmerControl(optimizer = "bobyqa",        # more stable than default Nelder_Mead
                             optCtrl   = list(maxfun = 1e5)))  # max function evaluations
```

### Save and load model

```{r}
#| echo: fenced
#| eval: false
saveRDS(lmer_gaze, here("output", "models", "lmer_gaze.rds"))
```

```{r}
#| echo: fenced
#| eval: true
lmer_gaze <- readRDS(here("output", "models", "lmer_gaze.rds"))
```


- `eval: false`: fitting and saving the model
- `eval: true`: loading in the saved model

::: {style="text-align: center; margin-top: 1em; color: white;"}
### ✏️ Exercises — 5 minutes {background-color="#8D203D"}

[Exercises 1: Analysis script](file:///C:/Users/Palleschi/OneDrive%20-%20Leibniz-ZAS/Documents/Workshops/publishr/_book/exercises/day2.html#data-and-code)

You will render a script that already follows these steps.
:::

## Reporting your model

- report model results inline using `broom.mixed::tidy()`
- extract the term of interest with `filter()`, then pull and format each parameter

````{r}
#| echo: true
lmer_tidy        <- broom.mixed::tidy(lmer_gaze, effects = "fixed", conf.int = TRUE)
b_lifetime_lmer  <- lmer_tidy |> filter(term == "lifetime1") |> pull(estimate)  |> round(2)
se_lifetime_lmer <- lmer_tidy |> filter(term == "lifetime1") |> pull(std.error) |> round(2)
t_lifetime_lmer  <- lmer_tidy |> filter(term == "lifetime1") |> pull(statistic) |> round(2)
p_lifetime_lmer  <- lmer_tidy |> filter(term == "lifetime1") |> pull(p.value)   |> fmt_p(3)
````

### Inline reporting

- embed extracted values in prose using inline `` `r` ``
- results update automatically when the model changes

````markdown
There was a main effect of lifetime, with longer first-pass reading times for dead 
versus living referents ($\beta$ = `r b_lifetime_lmer`, SE = `r se_lifetime_lmer`, 
*t* = `r t_lifetime_lmer`, *p* `r p_lifetime_lmer`).
````

> There was a main effect of lifetime, with longer first-pass reading times for dead 
versus living referents ($\beta$ = `r b_lifetime_lmer`, SE = `r se_lifetime_lmer`, 
*t* = `r t_lifetime_lmer`, *p* `r p_lifetime_lmer`).

