---
title: "Writing in Quarto"
subtitle: "Markdown, Citations, and Cross-References"
---
````{r}
#| output: false
pacman::p_load(tidyverse, here, gt, broom, glue)
````

````{r}
#| output: false
df   <- readRDS(here::here("data", "processed", "data_clean.rds"))
m1   <- readRDS(here::here("models", "m1.rds"))
tbl3 <- readRDS(here::here("output", "tables", "tbl3.rds"))
````

## Markdown basics

Quarto uses Markdown for text formatting. The most common formatting options are:

| Markdown | Output |
|---|---|
| `**bold**` | **bold** |
| `*italic*` | *italic* |
| `***bold italic***` | ***bold italic*** |
| `` `code` `` | `code` |
| `~~strikethrough~~` | ~~strikethrough~~ |
| `H~2~O` | H~2~O |
| `R^2^` | R^2^ |

### Headers

Use `##` for first-level sections within a chapter, `###` for subsections, and `####` for subsubsections. The chapter title is set in the YAML `title:` field and rendered as H1 automatically.

### Lists

Unordered lists use `-`, ordered lists use `1.`. Nested lists are indented with two spaces:
````markdown
- unordered item
  - nested item

1. ordered item
   1. nested item
````

### Block elements

Blockquotes use `>` and are useful for quoting stimuli or definitions:
````markdown
> The manager hired the assistant who had known the director.
````

Callout boxes draw attention to important information:
````markdown
::: callout-note
A note — good for asides and tips.
:::

::: callout-warning
A warning.
:::

::: callout-important
An important point.
:::
````

## Citations and APA formatting {#sec-citations}

### Setting up a bibliography

Add the following to your document YAML:
````yaml
bibliography: references.bib
csl: apa.csl
````

`references.bib` is a BibTeX file containing your references. `apa.csl` is a Citation Style Language file that controls formatting. Download `apa.csl` from the Zotero CSL repository:
````r
download.file(
  "https://www.zotero.org/styles/apa",
  here::here("apa.csl")
)
````

### Citation syntax

| Syntax | Output |
|---|---|
| `@Bates2015` | Bates et al. (2015) |
| `[@Bates2015]` | (Bates et al., 2015) |
| `[-@Bates2015]` | (2015) |
| `[@Bates2015, p. 12]` | (Bates et al., 2015, p. 12) |
| `[@Bates2015; @Gelman2020]` | (Bates et al., 2015; Gelman, 2020) |

In prose:
````markdown
Mixed-effects models were fitted following @Bates2015.
All analyses were conducted in R [@R2024].
````

### Getting BibTeX entries

The most reliable workflow is to manage references in Zotero and export as BibTeX. Alternatives include:

- **Google Scholar** — Cite → BibTeX → paste into `references.bib`
- **DOI** — use `doi2bib.org` to generate BibTeX from a DOI
- **R packages** — `toBibtex(citation("lme4"))` gives a citable BibTeX entry

### APA 7 formatting

With `csl: apa.csl` set in the YAML, Quarto handles citation formatting automatically. For PDF output in APA style, use `documentclass: apa7`:
````yaml
format:
  pdf:
    documentclass: apa7
````

Statistical symbols should be italicised in APA 7. Use `*...*` in Markdown — it renders correctly in both HTML and PDF:

| Statistic | Markdown | Output |
|---|---|---|
| Mean | `*M* = 310.8` | *M* = 310.8 |
| SD | `*SD* = 155.9` | *SD* = 155.9 |
| t-test | `*t*(48) = 2.31` | *t*(48) = 2.31 |
| p-value | `*p* < .001` | *p* < .001 |
| Cohen's d | `*d* = 0.45` | *d* = 0.45 |

## Cross-references {#sec-crossref}

Cross-references keep your figure, table, and section numbers in sync automatically — no more manually updating "see Table 3" after adding a table.

### Prefixes

Every cross-reference label must start with the correct prefix:

| Type | Prefix | Reference syntax |
|---|---|---|
| Figure | `fig-` | `@fig-mass` |
| Table | `tbl-` | `@tbl-desc` |
| Section | `sec-` | `@sec-results` |
| Equation | `eq-` | `@eq-model` |
| Example | `exm-` | `@exm-stimulus` |

### Figures

Add a `label` and `fig-cap` to the chunk options:
````markdown
```{{r}}
#| label: fig-rt
#| fig-cap: "Total reading time by condition."
ggplot(df, aes(x = lifetime, y = tt)) +
  geom_boxplot()
```
````

Then reference in prose:
````markdown
As shown in @fig-rt, reading times were longer for dead referents.
````

### Tables
````markdown
```{{r}}
#| label: tbl-desc
#| tbl-cap: "Descriptive statistics by condition."
tbl3 |> gt()
```
````
````markdown
Descriptive statistics are presented in @tbl-desc.
````

### Sections

Add a `{#sec-}` label to any heading:
````markdown
## Results {#sec-results}
## Discussion {#sec-discussion}
````
````markdown
We return to this point in @sec-discussion.
````

### Equations

Write LaTeX math in a display block and add a label:
````markdown
$$
y_i = \beta_0 + \beta_1 \text{Tense}_i + \beta_2 \text{Lifetime}_i + \varepsilon_i
$$ {#eq-model}
````
````markdown
The model is specified in @eq-model.
````

## Numbered example sentences {#sec-examples}

Pandoc's `(@)` syntax auto-numbers linguistic examples continuously across the document:
````markdown
(@) The manager hired the assistant.
(@) *The assistant was hired the manager.
````

Named examples can be cross-referenced in prose:
````markdown
(@cond-a) Der Dirigent hat den Komponisten *gekannt*, 
          der die Sinfonie geschrieben hat.
(@cond-b) Der Dirigent hat den Komponisten *gekannt*, 
          der gestorben ist.

The stimuli in (@cond-a) and (@cond-b) differed in the lifetime status
of the referent.
````

::: callout-tip
For sub-examples (1a, 1b), interlinear glosses, or full linguistic formatting in PDF output, use the `linguex` or `gb4e` LaTeX packages via `include-in-header` in your YAML. For HTML, a custom Lua filter is required.
:::

## Inline reporting {#sec-inline}

Inline R code embeds computed values directly in prose so they update automatically when data or models change.

### Basic inline reporting
````markdown
The sample comprised `r n_distinct(df$px)` participants.
The mean total reading time was `r round(mean(df$tt, na.rm = TRUE), 1)` ms
(*SD* = `r round(sd(df$tt, na.rm = TRUE), 1)`).
````

### Reporting model results

Define a helper function in `R/utils.R` and source it in your document:
````{r}
#| output: false
report_t <- function(model, term) {
  b   <- broom::tidy(model, conf.int = TRUE)
  row <- b[b$term == term, ]
  glue::glue(
    "*b* = {round(row$estimate, 2)}, ",
    "95% CI [{round(row$conf.low, 2)}, {round(row$conf.high, 2)}], ",
    "*t*({round(broom::glance(model)$df.residual)}) = {round(row$statistic, 2)}, ",
    "*p* {ifelse(row$p.value < .001, '< .001', paste0('= ', round(row$p.value, 3)))}"
  )
}
````

Use in prose:
````markdown
The effect of lifetime was significant (`r report_t(m1, "lifetimeDead")`).
````

::: callout-tip
Store `report_t()` and similar helpers in `R/utils.R` and source it at the top of your manuscript:
````r
source(here::here("R", "utils.R"))
````

This keeps your `.qmd` readable and makes the function reusable across documents.
:::

### Avoiding magic numbers

Any number that appears in your manuscript should come from your data or models — not typed manually. This prevents inconsistencies when data changes:
````markdown
<!-- Bad: typed manually -->
The sample comprised 48 participants.

<!-- Good: computed inline -->
The sample comprised `r n_distinct(df$px)` participants.
````

## Chapter slides
````{=html}
<a href="../slides/07_writing.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
````