---
title: "Writing in Quarto — Exercises"
format: html
bibliography: references.bib
csl: apa.csl
editor: source
---
````{r}
#| output: false
pacman::p_load(tidyverse, here, gt, broom, glue)
````
````{r}
#| output: false
df  <- readRDS(here::here("data", "processed", "data_clean.rds"))
m1  <- readRDS(here::here("models", "m1.rds"))
tbl3 <- readRDS(here::here("output", "tables", "tbl3.rds"))
````

## Exercise 1 — Cross-references

Produce a figure and a table below, then reference them in the paragraph underneath.
````{r}
#| label: fig-rt
#| fig-cap: "YOUR CAPTION HERE"

# YOUR PLOT CODE HERE
````
````{r}
#| label: tbl-desc
#| tbl-cap: "YOUR CAPTION HERE"

tbl3 |> gt()
````

As shown in ..., reading times differed by condition. Descriptive statistics are presented in ...

---

## Exercise 2 — Numbered examples

Write two example sentences from your stimuli and cross-reference them in prose.

(@) YOUR SENTENCE HERE
(@) YOUR SENTENCE HERE

The sentences ... illustrate ...

---

## Exercise 3 — Inline reporting

Complete the sentence below using inline R:

The sample comprised ... participants.
Models were fitted on log-transformed RTs (*M* = ..., *SD* = ...).
The effect of lifetime was significant (...).

---

## Exercise 4 — APA statistics

Rewrite the following in correct APA 7 format:

"The effect was significant (t = 2.31, p = 0.025, d = 0.45)."

YOUR VERSION: