---
title: "Tables — Exercises"
author: "Your Name"
format: html
editor: source
---

# Set up

## Packages

```{r}
#| output: false
library(tidyverse)
library(here)
library(gt)
library(kableExtra)
library(flextable)
library(conflicted)
```

Fix package conflicts

```{r}
conflict_prefer("footnote", "kableExtra")
conflict_prefer("filter", "dplyr")
```

```{r}
#| output: false
df <- readRDS(here::here("data", "processed", "exp1_clean.rds"))
```

## Data preparation

```{r}
# Table 3: total reading time by condition
tbl3 <- df |>
  filter(region == "verb") |>
  mutate(lifetime = factor(lifetime, levels = c("living", "dead"))) |>
  summarise(
    n      = n(),
    mean   = mean(tt,   na.rm = TRUE),
    sd     = sd(tt,     na.rm = TRUE),
    median = median(tt, na.rm = TRUE),
    iqr    = IQR(tt,    na.rm = TRUE),
    .by    = c(tense, lifetime)
  ) |>
  arrange(tense, lifetime)
```


```{r}
# Table 4: rating frequencies by condition
tbl4 <- df |>
  distinct(px, item, tense, lifetime, rating) |>
  mutate(lifetime = factor(lifetime, levels = c("living", "dead"))) |>
  summarise(n = n(), .by = c(tense, lifetime, rating)) |>
  mutate(
    total = sum(n),
    pct   = round(n / total * 100, 1),
    .by   = c(tense, lifetime)
  ) |>
  complete(tense, lifetime, rating = 1:7, fill = list(n = 0, pct = 0)) |>
  mutate(total = max(total, na.rm = TRUE), .by = c(tense, lifetime)) |>
  arrange(tense, lifetime, rating) |>
  mutate(n_pct = paste0(n, " (", pct, "%)")) |>
  select(-n, -pct) |>
  pivot_wider(names_from = rating, values_from = n_pct, names_prefix = "r")
```

---

## Exercise 1 — gt

Produce Table 4 (rating frequencies) using `gt`.  
Add a spanner label, bold headers, and a source note.  
Export and paste into the shared document.
```{r}
# YOUR CODE HERE
```

---

## Exercise 2 — kableExtra

Produce Table 4 (rating frequencies) using `kableExtra`.  
Add `add_header_above()`, `booktabs = TRUE`, and a footnote.  
Export to `.tex` and paste into Overleaf.
```{r}
# YOUR CODE HERE
```

---

## Exercise 3 — flextable

Produce Table 4 (rating frequencies) using `flextable`.  
Add a spanner row, bold headers, and a footer note.  
Export to Word and paste into the shared document.
```{r}
# YOUR CODE HERE
```
