---
editor: source
execute: 
  eval: false
  echo: fenced
---

# Session 2 Exercises {.unnumbered}

In the `publishr_workshop` RProject, open the `exercises/00-data_cleaning.qmd` 
Quarto script. Before you make any changes, render it.

## Data and Code

*Add appropriate headers below to keep your work structured.*

### Exercise 1: A simple code chunk

Add a code chunk below that computes `2 + 2` and renders it. Make sure the 
chunk is both printed (code visible) and evaluated (output visible).

### Exercise 2: Echo without eval

Change the chunk options on your Exercise 1 chunk so that the code is printed 
but not evaluated. Render and observe the difference.

::: callout-tip
Relevant chunk option: `eval: false`
:::

### Exercise 3: Silent chunk

Add a new code chunk that computes `10 * 10`. Set the chunk options so that 
neither the code nor the output appears in the rendered document. Render and 
verify it is invisible.

::: callout-tip
Relevant chunk option: `include: false`
:::

### Exercise 4: Load data

Add a code chunk that loads the sample dataset. Set the chunk options so that 
the code is not printed in the rendered output but the chunk is evaluated.

```{r}
df_raw <- readr::read_csv(here::here("data", "raw", "data_example.csv"))
```

::: callout-tip
Relevant chunk option: `echo: false`
:::

### Exercise 5: Anonymise data

Add a code chunk that anonymises the `px` value.

```{r}
set.seed(416) # for reproducibility

px_lookup <- df_raw |>
  distinct(px_code) |>                             # one row per participant
  mutate(px = sample(1:n(), n(), replace = FALSE)) # randomly assign 1:N

df_clean <- df_raw |>
  left_join(px_lookup, by = "px_code") |>
  select(-px_code) |>                              # remove original identifier
  relocate(px) |>
  arrange(px)
```

### Exercise 6: Save processed data

Save this anonymised data in `data/processed`.

```{r}
write_csv(df_clean, here::here("data", "processed", "data_clean.csv"))
```

**Assign code chunks accordingly!!!** Do you want to run all these steps each time you render, or is rendering a way to document a one-time process?

### Exercise 7: Summary statistics

Load in your saved data (what should `eval` be?).

Create a code chunk that filters to the verb region and computes the mean and 
SD of first-pass reading time (`gaze`), storing them as named objects. Set the 
chunk so it runs silently.

```{r}
m_gaze  <- # YOUR CODE HERE
sd_gaze <- # YOUR CODE HERE
```

::: callout-tip
You will need `filter(region == "verb")`, `mean(..., na.rm = TRUE)`, 
`sd(..., na.rm = TRUE)`, and `round(..., 1)`.
:::

### Exercise 8: Inline reporting

Report your values from Exercise 7 inline by completing the sentence below:

```
Mean first-pass reading time at the verb region was ___ ms (SD = ___ ms).
```

### Exercise 9: Global eval

Add the following to your YAML `execute:` block:

```yaml
execute:
  eval: false
```

Render. What happens? Which exercises break and why?

Now fix the problem by adding `#| eval: true` to the chunks that need to run 
(Exercises 4 and 5). Re-render and confirm everything works.

::: callout-note
This exercise illustrates why global chunk options need to be set carefully, 
and why it is often better to set `eval: false` only on specific chunks rather 
than globally.
:::

---

## Models

### Exercise 1: Analysis script

Open the script `analysis.qmd` and run each code chunk manually. Then render it.

### Exercise 2: Set up your manuscript script

Open `manuscript.qmd`. The global chunk option to hide code is already set. 
Render it to confirm it works.

Add inline code to fill in the following sentence (lines 137-139 in the Quarto script):

```
There was a main effect of lifetime, with longer first-pass reading times 
for dead versus living referents ($\beta$ = , SE = , 
*t* = , *p* ).
```

::: callout-tip
You will need `broom.mixed::tidy()` with `effects = "fixed"` and `conf.int = TRUE`, 
then `filter(term == "tense1")` and `pull()` for each value.
:::

### Exercise 3: Reporting other effects

So far our manuscript only reports effects of lifetime. Do the same for the Intercept and the effects of `tense1` and `trial_c`. How you structure the sentences and paragraph is up to you.

---

### Exercise 4: Fit a total reading time model

Switch back to `analysis.qmd`.

1. Filter `df_clean` to the verb region and apply contrast coding and trial 
centring as covered in the chapter, but this time use `tt` (total 
reading time) as your dependent variable instead of `gaze`. Exclude zeros 
as before.

2. Set sum contrast coding for `lifetime` and `tense`. Center `trial`.

2. Fit the following model:

```{r}
lmer_tt <- lmer(
  log(tt) ~ lifetime * tense + trial_c +
    (1 + lifetime | px) +
    (1 + tense | item),
  data    = df_verb,
  control = lmerControl(optimizer = "bobyqa",
                        optCtrl   = list(maxfun = 1e5))
)
```

3. Check for convergence warnings. Is the model singular?

4. Save the model:

```{r}
saveRDS(lmer_tt, here("output", "models", "lmer_tt.rds"))
```

---

### Exercise 5: Report the total reading time model in your manuscript

Switch to `manuscript.qmd`.

1. Load the new model:

```{r}
lmer_tt <- readRDS(here("output", "models", "lmer_tt.rds"))
```

2. Extract the values needed for a methods paragraph:

```{r}
n_obs_tt       <- nobs(lmer_tt)
n_px_tt        <- summary(lmer_tt)$ngrps[["px"]]
n_items_tt     <- summary(lmer_tt)$ngrps[["item"]]
formula_tt     <- deparse1(formula(lmer_tt))
optimizer_tt   <- lmer_tt@optinfo$optimizer
n_total_tt     <- n_px_tt * n_items_tt
n_dropped_tt   <- n_total_tt - n_obs_tt
pct_dropped_tt <- round(n_dropped_tt / n_total_tt * 100, 1)
```

3. Use these objects to fill in the blanks in `manuscript.qmd` under the 
Total reading times section.

4. Extract and report the lifetime main effect inline, following the same 
pattern as Exercise 2.

## References {.unnumbered}

::: {#refs}
:::