---
title: "Data Wrangling Recap"
subtitle: "Loading, Inspecting, and Cleaning Data"
editor_options: 
  chunk_output_type: console
---

```{r}
#| output: false
pacman::p_load(
  tidyverse,
  here,
  janitor
)
```

## Loading data

Use `readr` for CSV and `haven` for SPSS/Stata/SAS files. Always load from `data/raw/` and save processed outputs to `data/processed/`.

```{r}
# Example: use a built-in dataset as a stand-in
df <- read_csv(here("data", "raw", "exp1_data_tidy.csv"))
glimpse(df)
```

## Inspecting data

A quick structural overview with `head()` and `summary()`:

```{r}
summary(df)
```

```{r}
head(df, 10)
```


Check missingness per column:

```{r}
df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0)
```

## Cleaning with dplyr

A typical cleaning pipeline:

```{r}
df_clean <- df |>
  filter(type == "critical") |> 
  mutate(
    lifetime  = factor(lifetime, levels = c("living", "dead")),
    tense      = factor(tense, levels = c("PP", "SF"),
                      labels = c("Present Perfect", "Simple Future")),
    congruence      = factor(congruence, levels = c("+CON", "-CON"),
                      labels = c("Congruent", "Incongruent")),
    px     = factor(px),
    item = factor(item)
  ) |>
  rename(
    condition  = mm
  ) |> 
  select(px:rating,rt_value,bio, critical)

glimpse(df_clean)
```

::: callout-tip
Use `|>` (native pipe, R ≥ 4.1) rather than `%>%` for new code. It requires no package and is slightly faster.
:::

## Saving processed data

```{r}
#| eval: false
write_csv(df_clean, here::here("data", "processed", "exp1_clean.csv"))
```

Load it back in subsequent chapters with:

```{r}
#| eval: false
df_clean <- read_csv(here::here("data", "processed", "penguins_clean.csv"))
```

## Chapter slides

```{=html}
<a href="../slides/02_wrangling.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
```
