---
title: "Summary Statistics & Demographics Tables"
subtitle: "gtsummary, skimr, and friends"
---

```{r}
#| output: false
library(tidyverse)
library(gtsummary)
library(gt)
```

```{r}
#| output: false
df_clean <- palmerpenguins::penguins |>
  filter(!is.na(sex)) |>
  mutate(
    sex     = factor(sex, labels = c("Female", "Male")),
    species = factor(species)
  ) |>
  rename(
    bill_len = bill_length_mm,
    bill_dep = bill_depth_mm,
    flipper  = flipper_length_mm,
    mass     = body_mass_g
  )
```

## The demographics table

A demographics or "Table 1" is usually the first table in a methods section. It describes the sample: who was measured, on what variables, and how those break down by group.

## gtsummary

`gtsummary` produces publication-ready tables from data frames with minimal code and sensible defaults.

### Basic summary table

```{r}
df_clean |>
  select(species, sex, bill_len, bill_dep, flipper, mass) |>
  tbl_summary(
    by = sex,
    label = list(
      bill_len ~ "Bill length (mm)",
      bill_dep ~ "Bill depth (mm)",
      flipper  ~ "Flipper length (mm)",
      mass     ~ "Body mass (g)"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Variable**") |>
  modify_caption("**Table 1.** Sample characteristics by sex")
```

::: callout-note
`add_p()` automatically selects tests (t-test, Wilcoxon, chi-square) based on variable type. You can override with `test = list(...)`.
:::

### Stratified by multiple groups

```{r}
df_clean |>
  select(species, sex, mass) |>
  tbl_strata(
    strata = species,
    .tbl_fun = ~ .x |>
      tbl_summary(
        by = sex,
        statistic = all_continuous() ~ "{mean} ({sd})",
        digits = all_continuous() ~ 1
      )
  )
```

## Exporting gtsummary tables

For Word:

```{r}
#| eval: false
tbl <- df_clean |>
  select(species, sex, mass) |>
  tbl_summary(by = sex) |>
  as_flex_table()

flextable::save_as_docx(tbl, path = here::here("output", "tables", "table1.docx"))
```

For LaTeX:

```{r}
#| eval: false
tbl |>
  as_kable_extra(format = "latex", booktabs = TRUE) |>
  kableExtra::save_kable(here::here("output", "tables", "table1.tex"))
```

## Chapter slides

```{=html}
<a href="../slides/03_demographics.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
```
