---
title: "Output Formats"
subtitle: "HTML, Word, PDF, and LaTeX"
day: "Day 1"
---

::: {style="text-align: right; margin-bottom: 1em;"}
````{=html}
<a href="../slides/03_formats.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
````
:::


```{r}
#| echo: false
#| message: false
#| output: false

pacman::p_load(tidyverse, kableExtra)
if (!knitr::is_latex_output()) pacman::p_load(gt)
```

## Purpose

This chapter covers rendering a Quarto document to different output formats. The same `.qmd` source file can produce HTML, Word, PDF, and a LaTeX `.tex` file, but each format has specific YAML options and some practical differences worth understanding before committing to a workflow.

Rather than trying to support all formats simultaneously, the most pragmatic approach is to decide early which format best fits your collaboration context. The right choice depends on who your co-authors are, how they prefer to give feedback, and what the target journal requires. A workflow that is technically reproducible but creates friction for collaborators is unlikely to be sustainable in practice.

To illustrate: my own workflow has evolved considerably over time. Writing my dissertation entirely in R Markdown (with separate `.Rmd` files per chapter knitted together) worked well for a sole-author project but would have been impractical with collaborators. Sharing PDF output for feedback from co-authors on a journal article was workable but painful: comments arrived as annotations on a PDF, had to be manually transferred back to the source (by me), and any revision meant re-rendering and re-sharing (2/10, would not recommend). More recently, rendering a first draft from Rmarkdown to PDF, uploading the produced `.tex` document to Overleaf, and inviting collaborators to edit there directly has been a reasonable compromise, as it preserves the reproducible analysis pipeline while giving collaborators a familiar LaTeX editing environment (8/10: not perfect for reproducibility at the writing stage, but it facilitates collaboration in a way that a pure Quarto workflow currently cannot).

A further option is to collaborate in Google Docs, rendering results text and tables to Word and copying them across. This is admittedly imperfect from a reproducibility standpoint, as prose edits made in Google Docs are not tracked in the source `.qmd`. However, it can be a pragmatic workflow when collaborators are not willing or able to use anything else. Crucially, the analysis itself remains fully documented in R scripts and output files, even if the manuscript prose lives elsewhere. The results are still reproducible; only the writing layer is decoupled.

The sections below cover the main output formats, their YAML options, and the practical trade-offs involved.

## Rendering formats

Quarto separates the source document from its output format. The same `.qmd` file can be rendered to multiple formats by specifying them in the YAML `format:` block:

````yaml
format:
  html:
    embed-resources: true
  pdf:
    keep-tex: true
  docx:
    reference-doc: custom-reference.docx
````

Render all formats at once from the terminal:

````bash
quarto render manuscript.qmd
````

Or a single format:

````bash
quarto render manuscript.qmd --to docx
````

This separation of content from presentation is one of Quarto's core strengths: you write once and render to whatever format a collaborator, journal, or audience requires, without maintaining separate documents. However, I suggest you choose *one* output format until you have a specific need to render to different formats.

## HTML

HTML is the most flexible output format. It supports interactive elements, renders equations via MathJax, and can be hosted online (like this book, written in Quarto) or shared as a standalone file. For internal reports, supplementary materials, or workshop books, HTML is usually the best default.

````yaml
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: flatly
````

::: callout-note
## Self-contained HTML with `embed-resources`
By default, Quarto renders HTML with external dependencies (CSS, JavaScript, figures) stored in a separate folder alongside the `.html` file. This means sharing the file alone — by email, for example — will result in missing images and broken styling. Setting `embed-resources: true` bundles everything into a single self-contained `.html` file that can be shared without any accompanying folder.

The trade-off is file size and render time: `embed-resources: true` produces larger files and renders more slowly. For a book or website that will be hosted online and served from a web server, leave it as `false`.
:::

## Word

Word output is the most practical format for sharing with co-authors who are not using Quarto, and is widely accepted by journals for submission. The rendered `.docx` looks and behaves like any other Word document, co-authors can annotate, track changes, and comment without needing to know anything about Quarto.

````yaml
format:
  docx:
    reference-doc: custom-reference.docx
    toc: false
    number-sections: false
````

A reference `.docx` controls the visual styling of the output (fonts, margins, heading styles, and spacing) by mapping Quarto's document elements to Word styles. Generate a default one and then customise it:


````bash
quarto pandoc -o custom-reference.docx --print-default-data-file reference.docx
````

Open the generated file in Word, modify the paragraph and heading styles to match your journal's requirements, and point your YAML to it with `reference-doc: custom-reference.docx`.

::: callout-warning
## Citations in Word

Citations are rendered as plain formatted text in `.docx` output. They are not live fields linked to your `.bib` file, so editing them directly in Word — or adding new ones — will have no effect on the source document and will be overwritten on re-render. Always treat the `.qmd` as the source of truth, and re-render to update the `.docx` rather than editing it directly.

If co-authors need to edit citations in Word, the most practical solution is to use Zotero's Word plugin for the final submission copy, keeping the Quarto document as the reproducible record of the analysis.
:::

## PDF

PDF output is produced by rendering the `.qmd` to LaTeX and then compiling with a LaTeX engine (by default `pdflatex`). This gives access to the full power of LaTeX for typesetting, including precise control over layout, fonts, and mathematical notation. PDF is the appropriate format for journal submission when a `.pdf` or `.tex` file is required.

````yaml
format:
  pdf:
    documentclass: article   # or scrartcl, apa7
    papersize: a4
    fontsize: 12pt
    geometry: margin=2.5cm
    linestretch: 1.5
    number-sections: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{booktabs}    % professional table rules
        \usepackage{longtable}   % tables spanning multiple pages
        \usepackage{microtype}   % improved text justification
````

::: callout-tip
## keep-tex
Setting `keep-tex: true` saves the intermediate `.tex` file alongside the rendered PDF. This is useful in two situations: if the journal requires a raw `.tex` file for submission, or if you want to upload to Overleaf for final formatting tweaks or co-author editing. See @sec-apa7 for important caveats about citation handling in `.tex` output.
:::

## Format comparison

The table below summarises the key differences between output formats for common manuscript features.

````{r}
#| label: tbl-formats
#| tbl-cap: "Comparison of output formats for common manuscript features."
#| echo: false

check <- if (knitr::is_latex_output()) "yes"     else "✅"
cross <- if (knitr::is_latex_output()) "no"      else "❌"
warn  <- if (knitr::is_latex_output()) "limited" else "⚠️"

tbl_formats <- tibble::tribble(
  ~Feature,           ~HTML,                       ~PDF,                                  ~Word,
  "Citations",        paste(check, "from .bib"),   paste(check, "from .bib"),             paste(check, "rendered as text"),
  "Live .bib link",   check,                       paste(warn, "depends on cite-method"), cross,
  "Cross-references", check,                       check,                                 check,
  "Equations",        paste(check, "MathJax"),     paste(check, "native LaTeX"),          paste(warn, "limited"),
  "Table formatting", "`gt`",                      "`kableExtra`",                        "`flextable`",
  "Custom styles",    "CSS",                       "LaTeX",                               "reference .docx",
  "Self-contained",   "`embed-resources: true`",   "always",                              "always"
)

if (knitr::is_latex_output()) {
  tbl_formats |>
    kbl(booktabs = TRUE, format = "latex") |>
    kable_styling(latex_options = c("hold_position", "scale_down")) |>
    column_spec(1, width = "3cm") |>
    column_spec(2:4, width = "3.5cm")
} else {
  tbl_formats |>
    gt() |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
}
````

## APA 7 and `.tex` output {#sec-apa7}

The following considerations are only relevant if you intend to edit the `.tex` file after rendering — for example, uploading to Overleaf for final formatting or co-author editing. If you render directly to PDF, Word, or HTML from Quarto, citations will be formatted correctly by the default citeproc engine regardless of which option you choose. The issue is specifically about whether the `.tex` file contains live `\cite{}` commands that a LaTeX editor like Overleaf can resolve, or plain formatted text that it cannot.

For `.tex` editing workflows, there is no perfect solution. The three main options involve trade-offs between APA 7 accuracy and Overleaf compatibility:

**citeproc (default)**  
Uses `csl: apa.csl` to format citations at render time. Produces the most accurate APA 7 output of the three options, but resolves citations to plain text in the `.tex` file. Overleaf receives a fully formatted document but with no live citation commands — adding or editing references requires going back to the Quarto source.

**natbib + apacite**  
Passes citations through to LaTeX as live `\cite{}` commands, which Overleaf can resolve using your uploaded `.bib` file. However, `apacite` was written to APA 6th edition standards and has known formatting issues with APA 7 requirements, including author name truncation rules and DOI formatting.

**biblatex + biblatex-apa**  
The most APA 7-compliant LaTeX citation option, with live citations. Requires `biber` as the backend processor, which must be set manually in Overleaf's compiler settings. More complex to set up than the other options but the best choice if both APA 7 accuracy and live Overleaf citations are required.

````{r}
#| label: tbl-apa7
#| tbl-cap: "Citation backend options for .tex output."
#| echo: false

tbl_apa7 <- tibble::tribble(
  ~Option,                   ~`APA 7 accuracy`,    ~`Live .tex citations`, ~Overleaf,
  "citeproc + apa.csl",      paste(check, "best"), cross,                  paste(warn, "upload .tex only"),
  "natbib + apacite",        paste(warn, "APA 6"), check,                  paste(check, "with .bib"),
  "biblatex + biblatex-apa", paste(check, "good"), check,                  paste(check, "set biber in compiler")
)

if (knitr::is_latex_output()) {
  tbl_apa7 |>
    kbl(booktabs = TRUE, format = "latex") |>
    kable_styling(latex_options = c("hold_position", "scale_down"))
} else {
  tbl_apa7 |>
    gt() |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
}
````

Choose based on your priority: APA 7 accuracy or Overleaf compatibility. To switch from the default citeproc:

````yaml
# natbib
format:
  pdf:
    keep-tex: true
    cite-method: natbib
    include-in-header:
      text: |
        \usepackage[natbibapa]{apacite}

# biblatex
format:
  pdf:
    keep-tex: true
    cite-method: biblatex
    biblio-style: apa
````


## Chapter slides
````{=html}
<a href="../slides/08_formats.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
````