---
title: "Reporting Model Results"
subtitle: "modelsummary, broom, and inline reporting"
---

```{r}
#| output: false
library(tidyverse)
library(broom)
library(modelsummary)
```

```{r}
#| output: false
df_clean <- palmerpenguins::penguins |>
  filter(!is.na(sex), !is.na(body_mass_g)) |>
  mutate(
    sex     = factor(sex, labels = c("Female", "Male")),
    species = factor(species, levels = c("Adelie", "Chinstrap", "Gentoo"))
  )
```

## Fitting models

```{r}
m1 <- lm(body_mass_g ~ species, data = df_clean)
m2 <- lm(body_mass_g ~ species + sex, data = df_clean)
m3 <- lm(body_mass_g ~ species * sex, data = df_clean)
```

## broom: tidy model output

`broom::tidy()` returns model coefficients as a data frame — essential for inline reporting and custom tables.

```{r}
tidy(m2, conf.int = TRUE)
```

`broom::glance()` gives model-level fit statistics:

```{r}
glance(m2)
```

## Inline reporting

Access model values directly in prose using inline R code. For example:

```markdown
The effect of sex on body mass was b = `r round(tidy(m2)$estimate[4], 1)` g
(95% CI [`r round(tidy(m2, conf.int=TRUE)$conf.low[4], 1)`,
`r round(tidy(m2, conf.int=TRUE)$conf.high[4], 1)`]),
t(`r glance(m2)$df.residual`) = `r round(tidy(m2)$statistic[4], 2)`,
p `r ifelse(tidy(m2)$p.value[4] < .001, "< .001", paste0("= ", round(tidy(m2)$p.value[4], 3)))`.
```

::: callout-tip
Wrap repeated inline expressions in a small helper function stored in `R/utils.R` to keep prose readable.
:::

## modelsummary: regression tables

`modelsummary()` produces formatted regression tables from one or more models.

```{r}
modelsummary(
  list("Null" = m1, "+ Sex" = m2, "Interaction" = m3),
  stars     = TRUE,
  gof_map   = c("nobs", "r.squared", "adj.r.squared", "rmse"),
  coef_rename = c(
    "speciesChinstrap"      = "Chinstrap",
    "speciesGentoo"         = "Gentoo",
    "sexMale"               = "Male",
    "speciesChinstrap:sexMale" = "Chinstrap × Male",
    "speciesGentoo:sexMale"    = "Gentoo × Male"
  ),
  title = "Table 2. Linear models predicting body mass (g)"
)
```

## Exporting regression tables

For Word:

```{r}
#| eval: false
modelsummary(
  list("+ Sex" = m2),
  output = here::here("output", "tables", "regression.docx")
)
```

For LaTeX:

```{r}
#| eval: false
modelsummary(
  list("+ Sex" = m2),
  output  = here::here("output", "tables", "regression.tex"),
  booktabs = TRUE
)
```

## Chapter slides

```{=html}
<a href="../slides/05_models.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
```
