---
title: "Fitting and reporting models"
subtitle: "Fitting and reporting mixed models"
author: "Daniela Palleschi"
date: "2026-04-29"
date-format: "MMMM D, YYYY"
day: "Day 2"
editor_options: 
  chunk_output_type: console
---

::: {style="text-align: right; margin-bottom: 1em;"}
````{=html}
<a href="../slides/05_models.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
````
:::

## Purpose

This chapter covers fitting regression models to eyetracking data, extracting estimates for inline reporting, and producing publication-ready model summary tables, all within a reproducible Quarto document. We start with a simple linear model to establish the basics, then move to linear mixed-effects models to account for the repeated-measures structure of the data. We close with a Bayesian implementation of the same model using `brms`, and cover how to extract and report posterior estimates inline.

All models are fitted on first-pass reading time at the verb region from @palleschi_living_2025, with data available from the accompanying OSF repository [@palleschi_data_2026].

## Set-up {.unnumbered}

Load packages.

````{r}
#| output: false
#| message: false
pacman::p_load(here, tidyverse, lme4, lmerTest, brms, broom.mixed, gt, kableExtra)
````

Load data. N.B., we've skipped some filtering steps in our data preparation, so our results will be slightly different from the published results.

````{r}
df_clean <- read_csv(here("data", "processed", "data_clean.csv"))
````

Load in helper function to format p-values according to APA 7 style.

````{r}
source(here("R", "helpers.R"))
````

## Data preparation {#sec-data-prep}

Before fitting any model, we:

- filter to the region of interest (verb)
- remove any missing first-pass duration values (`> 0`)
- apply contrast coding to categorical predictors
- centre the trial order variable
- inspect distribution of our dependent variable

First, check our data is fully balanced.

````{r}
#| output: false
df_clean |> janitor::tabyl(px) |> count(n)
df_clean |> janitor::tabyl(item) |> count(n)
````

We see we have data from 24 participants with all of them contributing 480 observations, and 80 items all with 144 observations. This is a useful first step to catch any unexpected imbalance (missing participants, dropped trials, or items with too few observations) before the data enter a model.

Then filter so that we only have reading times from the `verb` region, and trials where first-pass reading time was greater than 0 (i.e., wasn't skipped):

````{r}
df_verb <- df_clean |>
  filter(region == "verb" & gaze > 0) 
````

::: {.callout-note collapse="true"}
#### Sanity check: observations per condition

A random slope can only be estimated if there are multiple observations per unit per condition cell; the more observations per cell, the more reliably slope variance can be distinguished from residual error [e.g., @barr_2013; @barr_et_al_2013]. A typical rule of thumb is N per cell > 5. Below this threshold, convergence issues with a maximal random effects structure are almost guaranteed in the frequentist framework (`lme4`, `lmerTest`). Bayesian models (`brms`) are more robust to sparse cells through partial pooling and prior regularisation, but low N per cell will still result in wide, uncertain posteriors for random effect parameters.

````{r}
library(patchwork)

df_verb |>
  count(item, lifetime, tense) |>
  count(lifetime, tense, n, name = "n_items")

df_verb |>
  count(px, lifetime, tense) |>
  count(lifetime, tense, n, name = "n_participants")
````

````{r}
#| echo: false
#| label: fig-obvs
#| fig-cap: "Number of observations per condition across participants (A) and items (B). For maximal model convergence N > 5 is the rule of thumb (but does not guarantee convergence)."
df_verb |>
  count(px, lifetime, tense) |>
  ggplot(aes(x = n, fill = interaction(lifetime, tense))) +
  geom_histogram(binwidth = 1, colour = "white", linewidth = 0.2) +
  labs(x = "N observations per participant", y = "N participants",
       fill = "Condition", title = "N first-pass > 0 per participant") +
  theme_bw() +
  df_verb |>
  count(item, lifetime, tense) |>
  ggplot(aes(x = n, fill = interaction(lifetime, tense))) +
  geom_histogram(binwidth = 1, colour = "white", linewidth = 0.2) +
  labs(x = "N observations per item", y = "N items",
       fill = "Condition", title = "N first-pass > 0 per item") +
  theme_bw() +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
````

Most participants contributed 20 observations per condition, though a small number contributed 18 or 16. Similarly, most items appeared 6 times per condition, with a small number appearing 5 or 4 times. The near-complete balance supports a maximal random effects structure for both grouping factors, but we may end up needing a simplified structure (especially given one item having 4 observations in the `living PP` condition).
:::

### Contrast coding {#sec-contrasts}

Both predictors are categorical with two levels and should be sum-coded so that the intercept represents the grand mean and each coefficient represents the deviation from it. We assign -0.5 to the theoretically "baseline" level in each case, creating a new variable with `_c` appended to the name to indicate it is centred:

- `lifetime`: `living` = -0.5, `dead` = +0.5
- `tense`: `pp` (present perfect) = -0.5, `sf` (simple future) = +0.5

````{r}
df_verb <- df_verb |>
  mutate(
    lifetime = factor(lifetime, levels = c("living", "dead")),
    tense    = factor(tense, levels = c("SF", "PP"))
  ) 

contrasts(df_verb$lifetime) <- c(-0.5, 0.5)
contrasts(df_verb$tense)    <- c(-0.5, 0.5)
````

Note you can achieve the same thing with `contr.sum(2) / 2`, which divides the default ±1 values by 2 to give ±0.5. Always check your contrasts after setting them:

````{r}
contrasts(df_verb$lifetime)
contrasts(df_verb$tense)
````

::: {.callout-note collapse="true"}
#### Alternative: `±0.5` coding with `if_else()`

The `if_else()` function creates explicit numeric columns directly in the data wrangling pipeline:

````{r}
#| eval: false
df_verb <- df_verb |>
  mutate(
    lifetime = if_else(lifetime == "living", -0.5, 0.5),
    tense    = if_else(tense == "PP", -0.5, 0.5)
  )
````

This achieves the same sum coding as `contr.sum(2) / 2` but with a few practical advantages: the coding is visible in the pipeline, it doesn't require the variable to be a factor, and it won't be lost if the data are re-read or re-factored. The tradeoff is that the original labels are overwritten, so plotting requires either a lookup or a separate label column.
:::

### Centre trial order

Critical trials were numbered 3–209, with 80 critical trials in total (the remainder being fillers). We centre within participant so that each participant's trial effect is expressed relative to the midpoint:

````{r}
df_verb <- df_verb |>
  mutate(trial_c = trial - mean(trial))
````

::: {.callout-note}
#### Why centre and sum-code predictors?

**Sum coding** (-0.5 / +0.5) means the model intercept is the grand mean of the response, and each coefficient is the difference between one level and the grand mean. This makes coefficients interpretable as main effects in the presence of an interaction, unlike treatment (dummy) coding where the intercept is the mean of the reference level only.

**Centring continuous predictors** (subtracting the mean) serves a similar purpose: it makes the intercept meaningful (the predicted value at the average level of the predictor) and reduces collinearity between main effects and interactions. For `trial_c`, we centre around the mean critical trial position. Because the trial sequence was pseudorandomised and held constant across participants, this is equivalent to subtracting a constant — no `group_by(px)` is needed. Both practices also improve model convergence in mixed-effects models.
:::

### Log-transform the dependent variable

Fixation duration measures are typically right-skewed: most fixations are short, but a long tail of unusually long fixations pulls the mean upward. This violates the normality assumption of linear models. A log-transformation compresses the upper tail and produces residuals that more closely approximate a normal distribution.

````{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-distribution
#| fig-cap: Positive skew in millisecond scale versus relative normality in log scale for first-fixation times at the verb region
#| fig-asp: .5
#| out-width: "80%"
library(patchwork)

df_verb |>
  ggplot(aes(x = gaze)) +
  geom_histogram(fill = "pink") +
  labs(title = "Millisecond scale") +
  df_verb |> 
  ggplot(aes(x = log(gaze))) +
  geom_histogram(fill = "lightblue") +
  labs(title = "Log scale") &
  theme_bw()
````

The data are slightly bimodal (two peaks), but this will be absorbed by the random effects.

::: {.callout-note}
#### Interpreting log-transformed estimates

Coefficients from a model fitted on `log(gaze)` are on the log-ms scale. To back-transform to the original millisecond scale, exponentiate:

````{r}
#| eval: false
exp(coef)       # back-transform a single coefficient
exp(conf.low)   # back-transform a confidence interval bound
````

A coefficient of 0.05 on the log scale means a multiplicative change of `exp(0.05)` $\approx$ 1.05, i.e. a 5% increase in fixation duration. For small effects, log-scale coefficients are approximately equal to proportional changes, but for larger effects exponentiation is necessary for accurate interpretation.
:::

## Linear mixed-effects model {#sec-lmer}

The appropriate model for repeated-measures data includes random effects for both participants and items. We use `lmer()` from the `lme4` package, with *p*-values provided by `lmerTest`.

### Random effects structure

Let's fit a model with the maximal random effects structure justified by our design and data:

```{r}
#| echo: true
#| eval: false
#| output: false
lmer_gaze_mm <- lmer(
  log(gaze) ~ lifetime * tense + trial_c +
    (1 + lifetime * tense | px) +
    (1 + lifetime * tense | item),
  data = df_verb
)
```


```{r}
#| eval: false
saveRDS(lmer_gaze_mm, here("output", "models", "lmer_gaze_mm.rds"))
```

We get the warning `boundary (singular) fit: see help('isSingular')`. This tells us our random effects structure is overparameterised. We'll skip how to find the model that converges and just run the model reported in the paper. I'm showing my code chunk options so you see my workflow:

```{r}
#| echo: fenced
#| eval: false
lmer_gaze <-
  lmer(log(gaze) ~ lifetime * tense + trial_c +
         (1 + lifetime | px) +
         (1 + tense | item), 
       data    = df_verb,
       control = lmerControl(optimizer = "bobyqa",        # more stable than default Nelder_Mead
                             optCtrl   = list(maxfun = 1e5)))  # max function evaluations
```

I run the models interactively, but set their code chunk options to `eval: false`. This is because I don't want to run every model when I render the document. For my final model, or any model that I want to interact with in the rendered document, I save it as an .rds file (also with `eval: false`):

```{r}
#| echo: fenced
#| eval: false
saveRDS(lmer_gaze, here("output", "models", "lmer_gaze.rds"))
```

Then I load in the same file immediately after this, with `eval: true`. This way when I render the document I can still e.g., print the summary of the model.

````{r}
#| echo: fenced
#| eval: true
lmer_gaze <- readRDS(here("output", "models", "lmer_gaze.rds"))
````

````{r}
summary(lmer_gaze)
````

::: {.callout-tip}
#### Convergence warnings

If the model fails to converge, the most common remedies in order of invasiveness are:

1. Switch optimiser: `lmerControl(optimizer = "bobyqa")` or `"Nelder_Mead"`
2. Increase the maximum number of function evaluations: `optCtrl = list(maxfun = 2e5)`
3. Remove the interaction from the random slopes: `(1 + lifetime + tense | px)`
4. Remove correlations between random effects: `(1 + lifetime * tense || px)`
5. Fall back to random intercepts only: `(1 | px) + (1 | item)`

In this case, the maximal by-item structure `(1 + lifetime * tense | item)` produced a singular fit, likely due to sparse cells (n = 6 per item per condition). Switching to `bobyqa`, increasing `maxfun`, and dropping some slopes resolved both issues. Always report what random effects structure was fitted and why, especially if you simplified from the maximal structure [@barr_et_al_2013; @matuschek_et_al_2017; @bates_et_al_2015].
:::

### Model summary table

We can use `broom.mixed::tidy()` to extract model output into a tidy dataframe.

```{r}
broom.mixed::tidy(lmer_gaze)
```


We can also pass the tidy table to `kbl()` from `kableExtra` for a formatted table, or `gt()` for HTML (kableExtra would also work, but gt() is better suited to HTML). You can do a lot of formatted to the table, and we also use our custom `fmt_p` function to format the *p*-values.

````{r}
#| label: tbl-lmer
#| code-fold: true
#| tbl-cap: "Linear mixed-effects model predicting first-pass reading time at the verb region. Fixed effects only."
tidy_tbl <- broom.mixed::tidy(lmer_gaze, effects = "fixed", conf.int = TRUE) |>
  mutate(
    across(c(estimate, std.error, conf.low, conf.high), \(x) round(x, 3)),
    statistic = round(statistic, 2),
    p.value   = fmt_p(p.value, table = TRUE)
  ) |>
  select(-effect)

if (knitr::is_latex_output()) {
  tidy_tbl |>
    mutate(term = str_replace_all(term, "_", "\\\\_")) |>
    kbl(booktabs = TRUE, format = "latex", escape = FALSE,
        col.names = c("Term", "$\\beta$", "SE", "$t$", "df", "$p$",
                      "95\\% CI lower", "95\\% CI upper")) |>
    kable_styling(latex_options = c("hold_position", "scale_down"))
} else {
  tidy_tbl |>
    gt() |>
    cols_label(
      term      = "Term",
      estimate  = "β",
      std.error = "SE",
      statistic = "t",
      df        = "df",
      p.value   = "p",
      conf.low  = "95% CI lower",
      conf.high = "95% CI upper"
    ) |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
}
````

### Inline reporting

````{r}
#| output: false
lmer_tidy     <- broom.mixed::tidy(lmer_gaze, effects = "fixed", conf.int = TRUE)
b_trial_lmer  <- lmer_tidy |> filter(term == "trial_c") |> pull(estimate)  |> round(2)
se_trial_lmer <- lmer_tidy |> filter(term == "trial_c") |> pull(std.error) |> round(2)
t_trial_lmer  <- lmer_tidy |> filter(term == "trial_c") |> pull(statistic) |> round(2)
p_trial_lmer  <- lmer_tidy |> filter(term == "trial_c") |> pull(p.value)   |> fmt_p(3)
````

````markdown
There was a main effect of lifetime, with longer first-pass reading times 
for dead versus living referents ($\beta$ = `r b_trial_lmer`, SE = `r se_trial_lmer`, 
*t* = `r t_trial_lmer`, *p* `r p_trial_lmer`).
````

> There was a main effect of lifetime, with longer first-pass reading times for dead versus living referents ($\beta$ = `r b_trial_lmer`, SE = `r se_trial_lmer`, *t* = `r t_trial_lmer`, *p* `r p_trial_lmer`).

::: {.callout-note}
#### Bimodality in the raw data is not necessarily a problem

Inspection of the raw first-pass reading times revealed a bimodal distribution in log space, reflecting a mixture of single-fixation and refixation trials. This might seem to violate the normality assumption of linear mixed-effects models, but the assumption applies to the *residuals*, not the raw data. Once the model has accounted for systematic variance (fixed effects) and between-participant and between-item variability (random effects), the residuals are expected to be approximately normally distributed even if the raw outcome variable is not.

Residual inspection confirms this: the residual histogram (@fig-residuals A) is unimodal and approximately symmetric, centred on zero, and the Q-Q plot (@fig-residuals B) shows good agreement with the theoretical normal distribution in the central range, with only minor deviation in the tails, which is typical for log-transformed reading time data. The residuals vs. fitted plot (@fig-residuals C) showed no systematic pattern across most of the fitted value range, with a slight upturn at higher fitted values indicating minor heteroscedasticity, which is common with reading time data even after log transformation and is unlikely to meaningfully affect inference.

````{r}
#| echo: false
#| label: fig-residuals
#| fig-cap: Residual plots for the linear mixed-effects model.
#| out-width: "80%"
#| fig-asp: .4
p_hist <- data.frame(r = resid(lmer_gaze)) |>
  ggplot(aes(x = r)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  labs(x = "Residuals", y = "Count", title = "Histogram of residuals")

p_qq <- data.frame(r = resid(lmer_gaze)) |>
  ggplot(aes(sample = r)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(x = "Theoretical quantiles", y = "Sample quantiles", title = "Q-Q plot")

p_resid <- data.frame(fitted = fitted(lmer_gaze), residual = resid(lmer_gaze)) |>
  ggplot(aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, colour = "red", linewidth = 0.7) +
  labs(x = "Fitted values", y = "Residuals", title = "Residuals vs fitted")

p_hist + p_qq + p_resid & theme_bw()
````

The practical takeaway is that residual plots, not plots of the raw outcome, are the appropriate diagnostic for assessing whether the normality assumption has been met.
:::

## Reporting the model in a methods section

Before reporting individual parameter estimates, a methods or results section should describe the model itself: the formula fitted, the data it was applied to, and any decisions made about the random effects structure. All of this can be extracted directly from the fitted model object and reported inline, ensuring consistency between the model and the write-up.

````{r}
#| output: false
lmer_gaze <- readRDS(here("output", "models", "lmer_gaze.rds"))

n_obs         <- nobs(lmer_gaze)
n_px          <- summary(lmer_gaze)$ngrps[["px"]]
n_items       <- summary(lmer_gaze)$ngrps[["item"]]
model_formula <- deparse1(formula(lmer_gaze))
optimizer     <- lmer_gaze@optinfo$optimizer
n_evals       <- format(lmer_gaze@optinfo$control$maxfun, scientific = FALSE, big.mark = ",")

# contrast coding
lifetime_contrasts_lmer <- contrasts(lmer_gaze@frame$lifetime)
tense_contrasts_lmer    <- contrasts(lmer_gaze@frame$tense)

# skipped trials
n_total    <- n_px * n_items
n_dropped  <- n_total - n_obs
pct_dropped <- round(n_dropped / n_total * 100, 1)
````

````markdown
The dataset comprised `r n_obs` observations from `r n_px` participants across 
`r n_items` items. `r n_dropped` trials (`r pct_dropped`%) were excluded due to 
first-pass reading times of zero, indicating the region was skipped. First-pass reading time at the verb region was analysed using a linear 
mixed-effects model fitted with `lme4` [@bates_et_al_2015] and `lmerTest` 
[@kuznetsova_et_al_2017], with Satterthwaite's method for degrees of freedom. 
The model formula was `r model_formula`. Categorical predictors were sum-coded 
(-0.5/+0.5): lifetime (`r rownames(lifetime_contrasts_lmer)[1]` = 
`r lifetime_contrasts_lmer[1]`, `r rownames(lifetime_contrasts_lmer)[2]` = 
`r lifetime_contrasts_lmer[2]`) and tense (`r rownames(tense_contrasts_lmer)[1]` 
= `r tense_contrasts_lmer[1]`, `r rownames(tense_contrasts_lmer)[2]` = 
`r tense_contrasts_lmer[2]`). Trial order was centred around the mean critical 
trial position. The by-item interaction slope was dropped from the maximal random 
effects structure due to a singular fit, likely reflecting sparse cells (n = 6 
observations per item per condition). The model was fitted using the `r optimizer` 
optimiser with a maximum of `r n_evals` function evaluations.
````

This will be printed as:

> The dataset comprised `r n_obs` observations from `r n_px` participants across 
`r n_items` items. `r n_dropped` trials (`r pct_dropped`%) were excluded due to 
first-pass reading times of zero, indicating the region was skipped.  First-pass reading time at the verb region was analysed using a linear mixed-effects model fitted with `lme4` [@bates_et_al_2015] and `lmerTest` [@kuznetsova_et_al_2017], with Satterthwaite's method for degrees of freedom. The model formula was `r model_formula`. Categorical predictors were sum-coded (-0.5/+0.5): lifetime (`r rownames(lifetime_contrasts_lmer)[1]` = `r lifetime_contrasts_lmer[1]`, `r rownames(lifetime_contrasts_lmer)[2]` = `r lifetime_contrasts_lmer[2]`) and tense (`r rownames(tense_contrasts_lmer)[1]` = `r tense_contrasts_lmer[1]`, `r rownames(tense_contrasts_lmer)[2]` = `r tense_contrasts_lmer[2]`). Trial order was centred around the mean critical trial position. The by-item interaction slope was dropped from the maximal random effects structure due to a singular fit, likely reflecting sparse cells (n = 6 observations per item per condition). The model was fitted using the `r optimizer` optimiser with a maximum of `r n_evals` function evaluations.

::: {.callout-tip}
#### What to report for a mixed-effects model

- The R packages used and their versions (extractable from `sessionInfo()`)
- The full model formula, including random effects structure and contrast coding scheme
- Any simplifications from the maximal structure and the reason (e.g. singular fit, convergence failure)
- The optimiser used and any non-default control parameters
- The number of observations, participants, and items
- The contrast coding scheme for categorical predictors
:::

## Bayesian mixed-effects model

We fit the same model using `brms`, which provides a full posterior distribution over all parameters rather than point estimates, giving us richer uncertainty quantification and a principled framework for incorporating prior knowledge. Importantly, we can also retain our maximal random effects structure because the priors regularise poorly-constrained parameters instead of producing the singular fits and convergence failures common in `lme4`. We use default priors for today's purposes, but in practice you would set and report weakly informative priors that reflect domain knowledge about plausible values for your dependent variable, and verify them with a prior predictive check.

### Fit the model

We use `lognormal()` as the likelihood, which is appropriate for right-skewed positive RT data. We run 2 chains with 2000 iterations (1000 warmup) — enough for a teaching example, but for a real model you'd want 4 chains and 4000 iterations. The `seed` argument ensures reproducibility.

````{r}
#| output: false
brm_gaze <- brm(
  gaze ~ lifetime * tense + trial_c +
    (1 + lifetime * tense | px) +  # by-participant
    (1 + lifetime * tense | item),       # by-item
  data   = df_verb,
  family = lognormal(),
  chains = 2,
  iter   = 2000,
  warmup = 1000,
  seed   = 416,
  # cores = parallel::detectCores() - 1  # parallelise chains
  file   = here("output", "models", "brm_gaze.rds")  # cache model
)
````

We get a low ESS warning because we're running a minimal number of chains and iterations for demonstration purposes. For a real model, change these settings to:

````{r}
#| eval: false
chains = 4,
iter   = 4000,
warmup = 2000,
````

Inspect model summary.

```{r}
summary(brm_gaze)
```


::: {.callout-tip}
#### Cache your models with `file =`

The `file` argument saves the fitted model to disk. On subsequent renders, `brms` will load the cached model rather than re-fitting, saving considerable time. The file is invalidated automatically if the model formula, data, or priors change. Use `here()` to keep the path relative to the project root.
:::

### Model diagnostics

Check convergence via trace plots and $\hat{R}$ values:

```{r}
plot(brm_gaze)      # trace plots and rank plots for each parameter
```

::: {.callout-note}
#### Interpreting convergence diagnostics

**Trace plots** show the sampled values for each parameter across iterations. A healthy trace plot looks like a "hairy caterpillar": the chains mix well, overlap completely, and show no trends or drifts. The posterior histogram alongside it should be smooth and unimodal.

**$\hat{R}$** (Rhat) measures agreement between chains. Values < 1.01 indicate convergence. **ESS** (effective sample size) measures how many independent samples you effectively have; both `Bulk_ESS` and `Tail_ESS` should exceed 1000. Low ESS means noisier estimates but not necessarily wrong ones; running more chains and iterations will fix it (recall that we cut our chains and iterations in half in the interest of running models faster during the workshop).

```{r}
posterior::summarise_draws(brm_gaze, "rhat", "ess_bulk", "ess_tail") |>
  summarise(
    max_rhat     = max(rhat,     na.rm = TRUE),
    min_ess_bulk = min(ess_bulk, na.rm = TRUE),
    min_ess_tail = min(ess_tail, na.rm = TRUE)
  )
```
:::

### Model summary table

````{r}
#| code-fold: true
#| label: tbl-brm
#| tbl-cap: "Bayesian mixed-effects model predicting first-pass reading time at the verb region. Posterior means, standard deviations, and 95% credible intervals for fixed effects."
brm_tidy_tbl <- as_draws_df(brm_gaze) |>
  select(starts_with("b_")) |>
  pivot_longer(everything(), names_to = "term", values_to = "value") |>
  group_by(term) |>
  summarise(
    mean  = round(mean(value), 3),
    sd    = round(sd(value), 3),
    q2.5  = round(quantile(value, 0.025), 3),
    q97.5 = round(quantile(value, 0.975), 3)
  ) |>
  mutate(term = str_remove(term, "^b_")) |>
  arrange(match(term, c("Intercept", "lifetime1", "tense1",
                        "lifetime1:tense1", "trial_c")))

if (knitr::is_latex_output()) {
  brm_tidy_tbl |>
    mutate(term = str_replace_all(term, "_", "\\\\_")) |>
    kbl(booktabs = TRUE, format = "latex", escape = FALSE,
        col.names = c("Term", "Posterior mean", "SD", "2.5\\%", "97.5\\%")) |>
    add_header_above(c(" " = 3, "95\\% Credible Interval" = 2), escape = FALSE)
} else {
  brm_tidy_tbl |>
    gt() |>
    cols_label(
      term  = "Term",
      mean  = "Posterior mean",
      sd    = "SD",
      q2.5  = "2.5%",
      q97.5 = "97.5%"
    ) |>
    tab_spanner(label = "95% Credible Interval", columns = c(q2.5, q97.5)) |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
}
````

### Posterior distributions

````{r}
#| code-fold: true
library(tidybayes)
library(ggdist)

brm_gaze |>
  gather_draws(b_lifetime1, b_tense1, `b_lifetime1:tense1`) |>
  mutate(.variable = factor(.variable,
    levels = c("b_lifetime1:tense1", "b_tense1", "b_lifetime1"),
    labels = c("Lifetime × Tense", "Tense", "Lifetime"))) |>
  ggplot(aes(x = .value, y = .variable, fill = .variable)) +
  stat_halfeye(
    point_interval = mean_qi,
    .width         = 0.95,
    alpha          = 0.8,
    height         = 0.6,
    normalize      = "groups"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("#7BAFD4", "#A8C87A", "#D4A8C7")) +
  guides(fill = "none") +
  labs(x = "Posterior estimate (log ms)", y = NULL,
       title = "Posterior distributions with 95% CrIs") +
  theme_classic(base_size = 12)
````

### Inline reporting

Extract posterior summaries for specific parameters:

```{r}
#| output: false
brm_draws <- as_draws_df(brm_gaze)

# lifetime main effect
b_life_mean  <- round(mean(brm_draws$b_lifetime1), 3)
b_life_sd    <- round(sd(brm_draws$b_lifetime1), 3)
b_life_lower <- round(quantile(brm_draws$b_lifetime1, 0.025), 3)
b_life_upper <- round(quantile(brm_draws$b_lifetime1, 0.975), 3)

# probability of direction: proportion of draws on the positive side
life_draws <- brm_draws$b_lifetime1
p_life_pos <- round(mean(life_draws > 0), 2)
```

Looking at the model summary above, we can see the lifetime effect has a 
posterior mean of 0.05 (Est.Error = 0.02, 95% CrI [0.00, 0.10]). The 
credible interval just excludes zero, suggesting weak evidence for longer 
first-pass reading times for dead versus living referents. We can extract 
and report this inline:

```markdown
There was a weak effect of lifetime, with longer first-pass reading times 
for dead versus living referents (posterior mean = `r b_life_mean`, 
SD = `r b_life_sd`, 95% CrI [`r b_life_lower`, `r b_life_upper`]; 
*P*($\beta$ > 0) = `r p_life_pos`).
```

Which renders as:

> There was a weak effect of lifetime, with longer first-pass reading times 
for dead versus living referents (posterior mean = `r b_life_mean`, 
SD = `r b_life_sd`, 95% CrI [`r b_life_lower`, `r b_life_upper`]; 
*P*($\beta$ > 0) = `r p_life_pos`).

::: {.callout-tip}
#### Reporting Bayesian results: key differences from frequentist

```{r}
#| echo: false
comparison_tbl <- tibble::tribble(
  ~concept,          ~lmer,                          ~brms,
  "Point estimate",  "$\\beta$ (fixed effect)",      "$\\beta$ (posterior mean/median)",
  "Uncertainty",     "Standard error (SE)",           "Posterior SD",
  "Interval",        "95\\% CI",                     "95\\% CrI",
  "Evidence",        "\\textit{p}-value",             "$P(\\beta > 0)$; CrI excludes zero"
)

if (knitr::is_latex_output()) {
  comparison_tbl |>
    kbl(booktabs = TRUE, format = "latex", escape = FALSE,
        col.names = c("", "Frequentist (\\texttt{lmer})", "Bayesian (\\texttt{brms})")) |>
    kable_styling(latex_options = "hold_position")
} else {
  tibble::tribble(
    ~` `,              ~`Frequentist (lmer)`,   ~`Bayesian (brms)`,
    "Point estimate",  "β (fixed effect)",       "Posterior mean or median",
    "Uncertainty",     "Standard error (SE)",    "Posterior SD",
    "Interval",        "95% CI (confidence)",    "95% CrI (credible)",
    "Evidence",        "*p*-value",              "*P*(β > 0); CrI excludes zero"
  ) |>
    gt() |>
    fmt_markdown(columns = everything()) |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
}
```

A few important distinctions in how you write up results:

- In `lmer`, $\beta$ is a **single point estimate**, meaning the model's best guess 
  for the effect, full stop. In `brms`, $\beta$ is a **distribution of 
  plausible values** (the posterior): the model returns not one number but a 
  range of values with associated probabilities. The mean or median of that 
  distribution is reported as the point estimate, but the full distribution 
  is what you actually work with for inference.
- The **95% CrI has a direct probability interpretation**: "there is a 95% 
  probability that the true effect lies between `r b_life_lower` and 
  `r b_life_upper`." A frequentist CI does not support this interpretation.
- **There is no *p*-value.** Instead, report the *probability of direction*: 
  the proportion of posterior draws on the expected side of zero. A value of 
  `r p_life_pos` means `r p_life_pos * 100`% of posterior draws are positive.
- **Do not say "significant."** Instead, describe the posterior directly: 
  "the credible interval excludes zero," "the bulk of the posterior mass is 
  positive," or "there is weak/strong evidence for an effect of X."
- The `Est.Error` in the `brms` summary is the **posterior SD** — not a 
  standard error in the frequentist sense. Label it *SD* in your write-up.
:::

## Reporting the Bayesian model in a methods section

The same information extracted for the frequentist model applies to the Bayesian model, with a few additions: the prior distributions, the sampler settings, and the key diagnostics used to verify convergence.

````{r}
#| output: false
m_brm <- readRDS(here("output", "models", "brm_gaze.rds"))

# sampler settings
n_chains  <- m_brm$fit@sim$chains
n_iter    <- m_brm$fit@sim$iter
n_warmup  <- m_brm$fit@sim$warmup
n_samples <- (n_iter - n_warmup) * n_chains

# number of observations, participants, items
n_obs_brm   <- nrow(m_brm$data)
n_px_brm    <- n_distinct(m_brm$data$px)
n_items_brm <- n_distinct(m_brm$data$item)

# contrast coding
lifetime_contrasts <- contrasts(m_brm$data$lifetime)
tense_contrasts    <- contrasts(m_brm$data$tense)

# model formula
brm_formula <- deparse1(m_brm$formula$formula)

# priors
brm_priors <- prior_summary(m_brm)
prior_intercept <- brm_priors |> filter(class == "Intercept") |> pull(prior)
prior_b         <- brm_priors |> filter(class == "b", coef == "") |> pull(prior)
prior_sd        <- brm_priors |> filter(class == "sd", coef == "") |> pull(prior) |> unique()
prior_sigma     <- brm_priors |> filter(class == "sigma") |> pull(prior)
prior_cor       <- brm_priors |> filter(class == "cor") |> pull(prior) |> unique()

# convergence
max_rhat <- round(max(rhat(m_brm), na.rm = TRUE), 3)
````

````markdown
First-pass reading time at the verb region was analysed using a Bayesian 
linear mixed-effects model fitted with `brms` [@burkner_2017], using the 
same formula as the frequentist model: `r brm_formula`. Lifetime was 
sum-coded with `r rownames(lifetime_contrasts)[1]` = 
`r lifetime_contrasts[1]` and `r rownames(lifetime_contrasts)[2]` = 
`r lifetime_contrasts[2]`; tense was sum-coded with 
`r rownames(tense_contrasts)[1]` = `r tense_contrasts[1]` and 
`r rownames(tense_contrasts)[2]` = `r tense_contrasts[2]`. The following 
priors were used: `r prior_intercept` for the intercept, `r prior_b` for 
fixed effects, `r prior_sd` for random effect standard deviations, 
`r prior_sigma` for the residual standard deviation, and `r prior_cor` for 
random effect correlations. The model was fitted with `r n_chains` chains 
of `r n_iter` iterations each, with `r n_warmup` warmup iterations, 
yielding `r n_samples` post-warmup samples. Convergence was assessed via 
trace plots and $\hat{R}$ values; all $\hat{R}$ values were below 1.01 
(max $\hat{R}$ = `r max_rhat`). The dataset comprised `r n_obs_brm` 
observations from `r n_px_brm` participants across `r n_items_brm` items.
````

> First-pass reading time at the verb region was analysed using a Bayesian 
linear mixed-effects model fitted with `brms` [@burkner_2017], using the 
maximal random effects structure: `r brm_formula`. Lifetime was 
sum-coded with `r rownames(lifetime_contrasts)[1]` = 
`r lifetime_contrasts[1]` and `r rownames(lifetime_contrasts)[2]` = 
`r lifetime_contrasts[2]`; tense was sum-coded with 
`r rownames(tense_contrasts)[1]` = `r tense_contrasts[1]` and 
`r rownames(tense_contrasts)[2]` = `r tense_contrasts[2]`. The following 
priors were used: `r prior_intercept` for the intercept, `r prior_b` for 
fixed effects, `r prior_sd` for random effect standard deviations, 
`r prior_sigma` for the residual standard deviation, and `r prior_cor` for 
random effect correlations. The model was fitted with `r n_chains` chains 
of `r n_iter` iterations each, with `r n_warmup` warmup iterations, 
yielding `r n_samples` post-warmup samples. Convergence was assessed via 
trace plots and $\hat{R}$ values; all $\hat{R}$ values were below 1.01 
(max $\hat{R}$ = `r max_rhat`). The dataset comprised `r n_obs_brm` 
observations from `r n_px_brm` participants across `r n_items_brm` items.

::: {.callout-note collapse="true"}
#### What to report: `lmer()` vs `brms()`

````{r}
#| label: tbl-reporting
#| tbl-cap: "Summary of what to report and how to extract it for frequentist and Bayesian mixed-effects models."
#| echo: false
reporting_tbl <- tibble::tribble(
  ~What,                   ~lmer,                                              ~brms,
  "Model formula",         "`deparse1(formula(m))`",                           "`deparse1(m$formula$formula)`",
  "N observations",        "`nobs(m)`",                                        "`nrow(m$data)`",
  "N participants",        "`summary(m)$ngrps[['px']]`",                       "`n_distinct(m$data$px)`",
  "N items",               "`summary(m)$ngrps[['item']]`",                     "`n_distinct(m$data$item)`",
  "Contrast coding",       "`contrasts(m@frame$lifetime)`",                    "`contrasts(m$data$lifetime)`",
  "Optimiser",             "`m@optinfo$optimizer`",                            "NUTS (default)",
  "Iterations",            "---",                                              "`m$fit@sim$iter`",
  "Warmup",                "---",                                              "`m$fit@sim$warmup`",
  "Chains",                "---",                                              "`m$fit@sim$chains`",
  "Convergence",           "`isSingular(m)`; check Hessian warnings",          "`rhat(m)`; all should be < 1.01",
  "Point estimate",        "`tidy(m) |> pull(estimate)`",                      "`as_draws_df(m) |> summarise(mean(b_...))`",
  "Uncertainty",           "SE: `tidy(m) |> pull(std.error)`",                 "SD: `as_draws_df(m) |> summarise(sd(b_...))`",
  "Interval",              "95% CI: `tidy(m, conf.int=TRUE)`",                 "95% CrI: `as_draws_df(m) |> quantile()`",
  "Test statistic",        "t: `tidy(m) |> pull(statistic)`",                  "P(b > 0): `mean(draws > 0)`",
  "p-value / evidence",    "p: `tidy(m) |> pull(p.value)`",                    "No p-value; report CrI + P(direction)",
  "Priors",                "---",                                              "`prior_summary(m)`",
  "Random effects",        "`as.data.frame(VarCorr(m))`",                      "`as_draws_df(m) |> select(starts_with('sd_'))`"
) |>
  mutate(What = paste0("**", What, "**"))

if (knitr::is_latex_output()) {
  reporting_tbl |>
    mutate(across(everything(),
      ~ str_replace_all(., "\\*\\*([^*]+)\\*\\*", "\\\\textbf{\\1}") |>
        str_replace_all("\\[\\[", "{[}{[}") |>
        str_replace_all("\\]\\]", "{]}{]}") |>
        str_replace_all(" \\|> ", " |> ") |>
        str_replace_all("`([^`]*) \\|> ([^`]*)`", "\\\\texttt{\\1 |> \\2}") |>
        str_replace_all("`([^`]+)`", "\\\\texttt{\\1}") |>
        str_replace_all("\\$", "\\\\$") |>
        str_replace_all("_", "\\\\_") |>
        str_replace_all("@", "\\\\@") |>
        str_replace_all("&", "\\\\&"))) |>
    kbl(booktabs = TRUE, format = "latex", escape = FALSE,
        col.names = c("What", "\\texttt{lmer()}", "\\texttt{brms()}")) |>
    kable_styling(latex_options = c("hold_position", "scale_down")) |>
    column_spec(1, width = "3cm") |>
    column_spec(2, width = "5.5cm") |>
    column_spec(3, width = "5.5cm")
} else {
  reporting_tbl |>
    gt() |>
    fmt_markdown(columns = everything()) |>
    cols_label(
      What = "**What**",
      lmer = md("**`lmer()`**"),
      brms = md("**`brms()`**")
    ) |>
    cols_width(
      What ~ px(180),
      lmer ~ px(250),
      brms ~ px(250)
    )
}
````
:::

::: {.callout-note}
#### Priors: what to set and what to report

Priors encode your assumptions about plausible parameter values *before seeing the data*. For eyetracking measures like first-pass reading time, we have strong domain knowledge: typical first-pass reading times fall in the 150–600 ms range, with a longer upper tail due to refixation trials. This justifies using weakly informative priors.

**What to report in a write-up:**

- The prior distributions for all parameters (intercept, slopes, random effect SDs, residual SD)
- A brief justification for each prior (domain knowledge, regularisation, or previous literature)
- Whether a **prior predictive check** was run to confirm priors generate plausible data before fitting

**Useful diagnostics to check and report:**

- `summary(brm_gaze)`: $\hat{R}$ values — all should be < 1.01
- `plot(brm_gaze)`: trace plots and rank plots for chain mixing
- `pp_check(brm_gaze)`: posterior predictive check — does the model reproduce the shape of your data?
- Effective sample size (`Bulk_ESS`, `Tail_ESS`) — should be > 1000

At minimum, a reproducible manuscript should report: prior distributions, $\hat{R}$ values, and a posterior predictive check.
:::

## Summary

````{r}
#| label: tbl-summary
#| tbl-cap: "Topics, key functions, and packages covered in this chapter."
#| echo: false
summary_tbl <- tibble::tribble(
  ~Topic,                         ~`Key functions`,                                      ~Package,
  "Contrast coding",              "`if_else()`, `mutate()`",                             "`dplyr`",
  "Centre predictors",            "`mutate()`, `group_by()`",                            "`dplyr`",
  "Sanity check",                 "`tabyl()`",                                           "`janitor`",
  "Simple linear model",          "`lm()`, `summary()`",                                 "base R",
  "Mixed-effects model",          "`lmer()`",                                            "`lme4`, `lmerTest`",
  "Bayesian model",               "`brm()`",                                             "`brms`",
  "Model summary table",          "`tidy()`, `gt()`",                                    "`broom`, `broom.mixed`, `gt`",
  "Extract estimates",            "`tidy()`, `filter()`, `pull()`",                      "`broom.mixed`, `dplyr`",
  "Posterior draws",              "`as_draws_df()`",                                     "`brms`",
  "Prior specification",          "`prior()`, `set_prior()`",                            "`brms`",
  "Model diagnostics",            "`pp_check()`, `plot()`, `summary()`",                 "`brms`",
  "Inline reporting",             "`` `r ...` ``",                                       "Quarto"
) |>
  mutate(Topic = paste0("**", Topic, "**"))

if (knitr::is_latex_output()) {
  summary_tbl |>
    mutate(
      `Key functions` = str_replace_all(`Key functions`,
                                        "`` `r \\.\\.\\.` ``",
                                        "\\\\texttt{r ...}"),
      across(everything(),
        ~ str_replace_all(., "\\*\\*([^*]+)\\*\\*", "\\\\textbf{\\1}") |>
          str_replace_all("`([^`]+)`", "\\\\texttt{\\1}") |>
          str_replace_all("\\$", "\\\\$") |>
          str_replace_all("_", "\\\\_") |>
          str_replace_all("@", "\\\\@"))) |>
    kbl(booktabs = TRUE, format = "latex", escape = FALSE) |>
    kable_styling(latex_options = c("hold_position", "scale_down")) |>
    column_spec(2, width = "5cm")
} else {
  summary_tbl |>
    gt() |>
    fmt_markdown(columns = everything()) |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    ) |>
    cols_width(
      Topic ~ px(200),
      `Key functions` ~ px(280)
    )
}
````

## Session Info {.unnumbered}

::: {.callout-important}
#### Reproducibility: Session Info

Always run `sessionInfo()` at the end of every script. This records your R version, package versions, and system information — an essential step for computational reproducibility.

````{r}
sessionInfo()
````

For more robust environment management, the `renv` package allows you to snapshot and restore the exact package versions used in a project.
:::