---
title: "Summary Tables for Publication"
subtitle: "gt and flextable"
---

```{r}
#| output: false
library(tidyverse)
library(gt)
library(flextable)
```

## Two tools, two targets

| Package     | Best for               |
|-------------|------------------------|
| `gt`        | HTML and PDF output    |
| `flextable` | Word (.docx) output    |

Both accept data frames and give fine-grained control over formatting.

## gt

`gt` tables are built in layers. Start with `gt()`, then add formatting functions.

```{r}
df_summary <- palmerpenguins::penguins |>
  filter(!is.na(sex), !is.na(body_mass_g)) |>
  group_by(species, sex) |>
  summarise(
    n        = n(),
    mean_mass = mean(body_mass_g),
    sd_mass   = sd(body_mass_g),
    .groups   = "drop"
  ) |>
  mutate(sex = str_to_title(sex))

df_summary |>
  gt(groupname_col = "species") |>
  cols_label(
    sex       = "Sex",
    n         = "N",
    mean_mass = "Mean (g)",
    sd_mass   = "SD (g)"
  ) |>
  fmt_number(columns = c(mean_mass, sd_mass), decimals = 1) |>
  tab_spanner(label = "Body mass", columns = c(mean_mass, sd_mass)) |>
  tab_header(
    title    = "Table 3",
    subtitle = "Body mass by species and sex"
  ) |>
  tab_source_note("Data: palmerpenguins") |>
  opt_row_striping()
```

### Exporting gt to LaTeX or Word

```{r}
#| eval: false
# Save as LaTeX
tbl_gt <- df_summary |> gt(groupname_col = "species")
gtsave(tbl_gt, here::here("output", "tables", "table3.tex"))

# Save as Word (via gt's docx support)
gtsave(tbl_gt, here::here("output", "tables", "table3.docx"))
```

## flextable

`flextable` gives pixel-level control for Word documents.

```{r}
df_summary |>
  rename(
    Species = species, Sex = sex, N = n,
    `Mean (g)` = mean_mass, `SD (g)` = sd_mass
  ) |>
  flextable() |>
  merge_v(j = "Species") |>
  valign(j = "Species", valign = "top") |>
  colformat_double(j = c("Mean (g)", "SD (g)"), digits = 1) |>
  bold(part = "header") |>
  autofit() |>
  set_caption("Table 3. Body mass by species and sex")
```

### Exporting flextable to Word

```{r}
#| eval: false
ft <- df_summary |>
  flextable() |>
  autofit()

save_as_docx(ft, path = here::here("output", "tables", "table3.docx"))
```

::: callout-note
For manuscripts that will be submitted as `.docx`, `flextable` is generally more reliable than `gt` because it maps directly to Word's native table model.
:::

## Chapter slides

```{=html}
<a href="../slides/06_tables.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
```
