---
title: "Tables"
---


````{r}
#| label: setup
#| output: false
#| message: false
pacman::p_load(here, tidyverse, kableExtra, reticulate)
if (!knitr::is_latex_output()) pacman::p_load(gt)
````

::: {.callout-note collapse="true"}
### Simple Markdown tables

For quick, small tables that do not need formatting, Markdown syntax is the simplest option:
````markdown
| Condition | Tense            | Lifetime |
|---|---|---|
| A         | Present Perfect  | Living   |
| B         | Present Perfect  | Dead     |
````

Markdown tables render in all output formats without any packages. However they offer no control over column widths, cell formatting, footnotes, or spanning headers. For anything that will appear in a manuscript, `gt` or `kableExtra` is preferable.
:::

### Which package to use

````{r}
#| label: tbl-packages
#| tbl-cap: "Recommended table packages by output format."
#| echo: false
tbl_packages <-
  tibble::tribble(
  ~Target,        ~Package,      ~`Key function`,
  "HTML",         "gt",          "gt()",
  "PDF / LaTeX",  "kableExtra",  "kbl(..., booktabs = TRUE)",
  "Word",         "gt",          "gtsave()"
) 

if (knitr::is_latex_output()) {
  tbl_packages |>
    kbl(booktabs = TRUE, format = "latex") |>
    kable_styling(latex_options = "hold_position")
} else {
  tbl_packages |>
    gt() |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )
}
````

## Tables from scratch with `tribble()`

Not all manuscript tables come from data. Methods sections, model specification tables, and condition overviews are often written out manually. `tibble::tribble()` lets you create a table row by row in a readable format, which can then be piped into `gt` or `kableExtra` for formatting.

````{r}
#| label: tbl-conditions
#| tbl-cap: "Experimental conditions."

tbl_conditions <-
  tibble::tribble(
  ~Condition, ~Tense,             ~Lifetime, ~Example,
  "A",        "Present Perfect",  "Living",  "Der Dirigent hat den Komponisten gekannt, der...",
  "B",        "Present Perfect",  "Dead",    "Der Dirigent hat den Komponisten gekannt, der...",
  "C",        "Simple Past",      "Living",  "Der Dirigent kannte den Komponisten, der...",
  "D",        "Simple Past",      "Dead",    "Der Dirigent kannte den Komponisten, der..."
)

if (knitr::is_latex_output()) {
  tbl_conditions |>
    kbl(
      format    = "latex",
      booktabs  = TRUE,
      col.names = c("Condition", "Tense", "Lifetime", "Example stimulus")
    ) |>
    kable_styling(latex_options = c("hold_position", "scale_down")) |>
    add_header_above(c(" " = 1, "Condition factors" = 2, " " = 1)) |>
    column_spec(4, width = "6cm") |>
    kableExtra::footnote(
      general           = "Only the critical verb region is shown.",
      general_title     = "Note.",
      footnote_as_chunk = TRUE
    )
} else {
  tbl_conditions |>
    gt() |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    ) |>
    tab_spanner(
      label   = "Condition factors",
      columns = c(Tense, Lifetime)
    ) |>
    cols_label(Example = "Example stimulus") |>
    cols_width(Example ~ px(300)) |>
    tab_source_note("Only the critical verb region is shown.")
}
````

## Cross-referencing tables

To cross-reference a table in prose, add a `label` starting with `tbl-` and a `tbl-cap` to the chunk options:
````markdown
```{{r}}
#| label: tbl-desc
#| tbl-cap: "Total reading time (ms) by condition."
tbl_desc |> gt()
```
````

Then reference in prose with `@tbl-desc`:
````markdown
Descriptive statistics are presented in @tbl-desc.
````

This renders as "Descriptive statistics are presented in Table 1." and updates automatically if tables are added or reordered.

::: callout-warning
Cross-references only work when the chunk label begins with `tbl-`. A label like `#| label: desc-table` will not be recognised as a table reference.
:::

## Summary tables

Summary tables in manuscripts serve two purposes: describing the sample (demographics, condition counts) and reporting descriptive statistics (means, SDs, reading times). The appropriate package depends on your target output format.



### Preparing summary data

````{r}
#| output: false
#| eval: false
df_clean <- readRDS(here::here("data", "processed", "exp1_clean.rds"))

tbl_desc <- df_clean |>
  filter(region == "verb") |>
  mutate(lifetime = factor(lifetime, levels = c("living", "dead"))) |>
  summarise(
    n      = n(),
    mean   = mean(tt,   na.rm = TRUE),
    sd     = sd(tt,     na.rm = TRUE),
    median = median(tt, na.rm = TRUE),
    iqr    = IQR(tt,    na.rm = TRUE),
    .by    = c(tense, lifetime)
  ) |>
  arrange(tense, lifetime)
````

### Key functions

````{r}
#| label: tbl-gt-functions
#| tbl-cap: "Key gt functions for manuscript tables."
tbl_gt_func <-
  tibble::tribble(
  ~Function,          ~Purpose,
  "tab_header()",     "Add title and subtitle",
  "tab_spanner()",    "Group columns under a label",
  "cols_label()",     "Rename columns",
  "fmt_number()",     "Format numeric columns",
  "tab_style()",      "Apply cell styles",
  "tab_source_note()", "Add a general note",
  "tab_footnote()",   "Add a footnote to a specific cell",
  "gtsave()",         "Export to HTML, PDF, or Word")

  
if (knitr::is_latex_output()) {
  tbl_gt_func |>
    kbl(
      booktabs = TRUE,
      format   = "latex",
      caption  = "Key gt functions for manuscript tables."
    ) |>
    kable_styling(latex_options = c("hold_position", "scale_down")) |>
    column_spec(2, width = "8cm")
} else {
  tbl_gt_func |>
    gt() |>
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    ) |>
    tab_header(title = "Key gt functions") |>
    cols_width(Purpose ~ px(350))
}
````

````{r}
#| label: tbl-kbl-functions
#| tbl-cap: "Key kableExtra functions for manuscript tables."
tbl_kbl_func <-
  tibble::tribble(
  ~Function,              ~Purpose,
  "kbl()",                "Initialise table",
  "kable_styling()",      "Set LaTeX or HTML styling options",
  "add_header_above()",   "Span columns under a label",
  "column_spec()",        "Format specific columns",
  "collapse_rows()",      "Merge repeated row values",
  "kableExtra::footnote()", "Add general or labelled notes",
  "save_kable()",         "Export to .tex, .html, or .pdf"
)

if (knitr::is_latex_output()) {
    tbl_kbl_func |>
      kbl(booktabs = TRUE, format = "latex") |>
      kable_styling(latex_options = c("hold_position", "scale_down")) |>
      column_spec(2, width = "8cm")
  } else {
    tbl_kbl_func |>
      gt() |>
      tab_style(
        style     = cell_text(weight = "bold"),
        locations = cells_column_labels()
      ) |>
      cols_width(Purpose ~ px(350))
  }
````

### Exporting tables

````{r}
#| eval: false
# gt to Word or PDF
tbl_gt <- tbl_desc |> gt()
gtsave(tbl_gt, here::here("output", "tables", "tab_desc.docx"))
gtsave(tbl_gt, here::here("output", "tables", "tab_desc.pdf"))

# kableExtra to LaTeX
tbl_desc |>
  kbl(booktabs = TRUE, digits = 1) |>
  kable_styling() |>
  save_kable(here::here("output", "tables", "tab_desc.tex"))
````
