---
title: "Exporting to LaTeX & Word"
subtitle: "Format-specific YAML, templates, and common fixes"
---

## Two targets, two strategies

The same Quarto source file can render to both PDF (via LaTeX) and Word — but each format has quirks. This chapter covers the most common issues and how to resolve them.

## PDF via LaTeX

### YAML for journal-style PDFs

```yaml
format:
  pdf:
    documentclass: article   # or apa7, scrartcl, etc.
    papersize: a4
    fontsize: 12pt
    geometry: margin=2.5cm
    linestretch: 1.5
    number-sections: true
    colorlinks: true
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{microtype}
        \usepackage[natbibapa]{apacite}
```

::: callout-warning
`natbib` and `biblatex` conflict — use one or the other. For APA style, `\usepackage[natbibapa]{apacite}` with `natbib: true` in YAML is reliable.
:::

### Tables in LaTeX

`kableExtra` is the most stable option for LaTeX tables:

```{r}
#| eval: false
library(kableExtra)

df_summary |>
  kbl(
    format   = "latex",
    booktabs = TRUE,
    caption  = "Body mass by species and sex",
    digits   = 1
  ) |>
  kable_styling(latex_options = c("striped", "hold_position")) |>
  collapse_rows(columns = 1, latex_hline = "major")
```

### Figures in LaTeX

Always export as `cairo_pdf` and reference from Quarto:

```{r}
#| eval: false
ggsave(
  "output/figures/fig1.pdf",
  plot   = last_plot(),
  device = cairo_pdf,
  width  = 14, height = 9, units = "cm"
)
```

Then include in `.qmd`:

```markdown
![Body mass by species.](output/figures/fig1.pdf){#fig-mass width=80%}
```

## Word (.docx)

### Reference documents

A reference `.docx` lets you pre-define styles (heading levels, body text, margins). Create one in Word, then point Quarto to it:

```yaml
format:
  docx:
    reference-doc: custom-reference.docx
```

To generate a starting reference document: `quarto pandoc -o custom-reference.docx --print-default-data-file reference.docx`

### Tables in Word

Use `flextable` — it maps directly to Word's table model and respects heading styles:

```{r}
#| eval: false
library(flextable)

ft <- df_summary |>
  flextable() |>
  autofit() |>
  bold(part = "header") |>
  set_caption("Body mass by species and sex")

save_as_docx(ft, path = "output/tables/table1.docx")
```

Or insert inline in Quarto and let it render into the document directly:

```{r}
#| eval: false
#| tbl-cap: "Body mass by species and sex."
#| label: tbl-mass
df_summary |>
  flextable() |>
  autofit()
```

### Figures in Word

High-DPI PNG works best for Word:

```{r}
#| eval: false
ggsave(
  "output/figures/fig1.png",
  plot  = last_plot(),
  width = 14, height = 9, units = "cm",
  dpi   = 300
)
```

Quarto includes figures inline automatically when rendering to `.docx`.

## Rendering both formats

From the terminal:

```bash
quarto render manuscript.qmd --to pdf
quarto render manuscript.qmd --to docx
```

Or add both to the YAML `format:` block and run `quarto render manuscript.qmd` to produce both at once.

## Common problems and fixes

| Problem | Fix |
|---|---|
| Missing LaTeX package | Add `\usepackage{...}` to `include-in-header` |
| Table overflows page | Use `latex_options = "scale_down"` in `kableExtra` |
| Font not found in PDF | Use `cairo_pdf` device; set `mainfont` in YAML |
| Bibliography not rendering | Check `bibliography:` path and CSL file location |
| Word styles not applying | Regenerate reference.docx and re-apply styles |

## Chapter slides

```{=html}
<a href="../slides/08_exporting.html" class="btn btn-outline-primary" target="_blank">
  Open slides ↗
</a>
```
